---
title: "Tutorial for Week 2"
output: 
  learnr::tutorial:
  progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(eval = FALSE)
```

## Welcome

In this tutorial, we will continue our introduction to R by exploring the topic of vectors. Understanding vectors is essential for many of the tasks you will be performing in R. We will also cover how to perform mathematical operations and other important functions with vectors.

When you are ready, click below to get started.

## Topic 1: Vectors

**Vectors** are an important building block of databases and data frames. A vector is a type of object, just like we created before, but it involves **MULTIPLE** elements. Vectors are very useful because they can end up serving as either a row or a column of a database.

Vectors can contain numbers, characters, and other data types we will learn about.

### The `c()` command

An essential command to create vectors is the `c()` command, which stands for "combine" or "concatenate". `c()` does just that -- it allows us to combine elements to create a vector. This is the first command that we will use that takes multiple "arguments" or elements that go within the parentheses. With `c()`, you place in the parentheses all of the elements you wish to combine into one vector, and separate each element using a comma. 

For example, the below chunk of code uses the `c()` command to combine four numbers to create a vector called `first_vector`.

```{r, first-vector, eval = TRUE, exercise = TRUE}

first_vector <- c(1, 5, 6, 8) # create the vector
first_vector # return the vector to see output and ensure it worked

```


Note that vectors can also combine non-numerical elements. As we did before, when we are dealing with characters, we have to use `" "` to make sure R knows these are character elements. For example:

```{r, first-vector2, eval = TRUE, exercise = TRUE}

second_vector <- c("I", "love", "political", "science") # create the vector

second_vector #return the vector

```


Now you practice! Create your own numerical vector called `my_vector` and give it 4 numbers of your choice.

```{r vec_ex1, exercise = TRUE}

```
```{r vec_ex1-hint}
my_vector <- c(__, __, __, __) #include four numbers of your choosing
```


Now do this again, but create a vector called `my_name` and fill it with your name. Give it one element for each of your names. For example, if you name is Lucy Dacus, you would have one element for "Lucy" and one for "Dacus". But, if your name is longer -- maybe Whitney Elizabeth Houston -- you would have one element for "Whitney", one for "Elizabeth", and one for "Houston". If your name is Madonna, it would just be "Madonna".

```{r vec_ex2, exercise = TRUE}

```
```{r vec_ex2-hint-1}
my_name <- c(___, ___) 
```
```{r vec_ex2-hint-2}
my_name <- c("Your", "Name", "Here")
```


## Topic 2: Math with Vectors

Now that we are able to create vectors, we can also do math with vectors. The below video will walk you through how we can perform mathematical operations with vectors.


<iframe src="https://drive.google.com/file/d/1MjwJM-HRMPg6D8hPzMh3aQKInDeJyPga/preview" width="900" height="490" allow="autoplay"></iframe>


Below, we'll walk through that example again and you can execute the provided code.

First, we'll take a look at our two vectors with information on the number of protests that occurred across three towns in 2015 and 2016. I want to know the total number of protests in each town for these two years.


```{r protest, exercise = TRUE}
# create my vectors
protest_15 <- c(5, 10, 8)

protest_16 <- c(2, 7, 3)

```

With this information, we can then add these two vectors together:

```{r protest_add, exercise = TRUE, exercise.setup = "protest"}

protest_15 + protest_16

```

Note that we can also save the result of this addition as its own vector:

```{r protest_add2, exercise = TRUE, exercise.setup = "protest"}

sum_protest <- protest_15 + protest_16 # save the object

sum_protest # return the result

```

### Using Math to Summarize Vectors

So far we have used math on vectors to create a whole new ***vector***.  Now we're going to see some math commands that you can perform on vectors to produce a ***single number***. R also has a number of built-in functions that will help us with this task. Two of them will be especially helpful as you move forward. The first function, mean(), calculates the arithmetic mean of a vector. The second function, sum(), calculates the sum of a vector. To see how these functions work, let's compute the mean and the sum of our `protest_15` vector.

```{r protest_add3, exercise = TRUE, exercise.setup = "protest"}

mean(protest_15) # calculate the mean

sum(protest_15) # calculat the sum 
```

### Practice

Your turn! Now you will be tasked with creating two vectors and performing some calculations with them. Let's imagine you have information about two political science classes at UNC. You have the percentage of students present in each of these classes from Monday --> Friday. 
Create two vectors with the following information:

* `class1` - Mon. 80%, Tues 95%, Weds.90%, Thurs 98%, Fri 88%
* `class2` - Mon. 82%, Tues 96%, Weds. 88%, Thurs 90%, Fri 80%

NOTE: Do not include the `%` sign when you are creating these vectors. Just type the number corresponding to the percent (e.g., 88 for 88%).

Calculate the difference between the attendance percentages of Class 1 and Class 2 for each day of the week by subtracting the Class 2 percentage from the Class 1 percentage. Save this in an object called `class_diff`.

Below, some initial code has been provided to you to help you out. Use this code as a starting place to complete this task.
 

```{r class_ex, exercise = TRUE, eval = F}
# create two vectors first
class1 <-
  
class2 <-


# then calculate the difference for each day

____ <- _____ - ______

```

```{r class_ex-hint-1}

# start with this 
class1 <- c(______)
class2 <- c(______)

```

```{r class_ex-hint-2}

# start with this 
class1 <- c(______)
class2 <- c(______)

# then subtract

class_diff <- _______-

```
```{r class_ex-hint-3}

# start with this 
class1 <- c(80, 95, 90, 98, 88)
class2 <- c(82, 96, 88, 90, 80)

# then subtract

class_diff <- class1 - class2

```

## Topic 3: Naming Vectors

Now that we have objects with multiple elements, it can also be useful to name those elements. This can help us keep track of the information a little bit better. For example, let's return to the example of protest. We are tracking the number of protests across three towns/cities in the years 2015 and 2016. But before, we didn't label those cities by name. But, we can do this, using the `names()` function. See below to see how it works


```{r, protest1, eval = TRUE, exercise = TRUE}
# first create my vectors
protest_15 <- c(5, 10, 8)
protest_16 <- c(2, 7, 3)

# now, use the names function to give these two objects names
names(protest_15) <- c("Chapel Hill", "Durham", "Raleigh")
names(protest_16) <- c("Chapel Hill", "Durham", "Raleigh")


#create a new object and add them together
total_protest <- protest_15 + protest_16

#return that object - what do you see with the numbers now?
total_protest
```

When you now return the object, you will see that each element of the vector will have these names - Chapel Hill, Durham, and Raleigh. 

Since we have two objects which we are giving the same names, we can also save those names in a separate character vector, and do the following:

```{r, protest2, eval = TRUE, exercise = TRUE}
# first create my vectors
protest_15 <- c(5, 10, 8)
protest_16 <- c(2, 7, 3)

# now, use the names function to give these two objects names
cities <- c("Chapel Hill", "Durham", "Raleigh")

# take the vector protest_15, and assign each element a name based on whatever text is in the vector called cities
names(protest_15) <- cities

# now do the same for the protest_16 vector
names(protest_16) <- cities


#create a new object and add them together
total_protest <- protest_15 + protest_16

#return that object - what do you see with the numbers now?
total_protest
```


Now you try naming a vector! Create a vector corresponding to the number of NCAA men's basketball wins with the following information. Store the number of wins in a vector called `wins_men`. Then name each element of the vector with the appropriate school's name.

* UNC - 6 wins
* Duke - 5 wins
* NC State - 2 wins

```{r bball_ex, exercise = TRUE}
# create the vector
wins_men <- _____


# then name the vector

names(_____) <- _________

```
```{r bball_ex-hint-1}

# create the vector
wins_men <- c(______)


# then name the vector

names(____) <- c(_______)

```
```{r bball_ex-hint-2}

# create the vector
wins_men <- c(6, 5, 2)


# then name the vector

names(wins_men) <- c("UNC", "Duke", "NC State")

```



## Topic 4: Selecting an Element of a Vector

Vectors are objects which store multiple elements of data. While this is great, it is important that we are still able to select certain elements, or parts, of those vectors. We will learn about this more in depth in class, but will provide an introduction and some practice here. 

The most basic way to select an element is by what we will call “indexing” a vector. To do this in R, you can put square brackets `[]` after the name of a vector. Then inside the brackets, you can select a particular element by putting either:

1. The name of that element
2. The physical place of an element in a vector (e.g., it is the first or second element)


So, for example, let's revisit that vector you created with the number of men's basketball NCAA championships. Say we want to isolate how many wins UNC had. We can select that element using square brackets, or `[]`. See below

```{r mens_bball, exercise=TRUE, eval = TRUE}
# create the vector
wins_men <- c(6, 5, 2)
# then name the vector
names(wins_men) <- c("UNC", "Duke", "NC State")


# select UNC's wins by name
wins_men["UNC"]


# select UNC's wins by element number
wins_men[1]

```

Let's look at what the code did above. It created the vector `wins_men` and named the elements of the vector using `name()`. We then indexed to select just the UNC information in two ways:

1. By using the name of the element we wanted: `wins_men[“UNC”]`
2. By using the fact that UNC is the first element of the vector: `wins_men[1]`


You can also select multiple elements! If you want to do this, you will have to use the `c()` command within the `[]`. See below for how we would select *both* UNC and Duke wins.


```{r mens_bball2, exercise=TRUE, eval = TRUE, exercise.setup = "mens_bball"}

# select UNC and Duke's wins

wins_men[c("UNC", "Duke")]

# or

wins_men[c(1, 2)]
```



Typically, it is recommended that you select a vector element by a name. That is because you sometimes will be dealing with vectors that have so many elements that it is unreasonable to know where exactly one element is located. However, it is important to know how to select by number as well, as this will come in handy.

### Practice Selecting Elements

Now, you try! We will now take a look at the NCAA women's basketball teams that have won the most championships. Create a new vector called `wins_women` which has the number of wins for the following schools.

* UConn - 11
* Tennessee - 8
* Baylor - 3
* Stanford - 3
* South Carolina - 2
* Notre Dame - 2

Name this vector such that each element corresponds to its school name. Then, using `[]`, show the following information:

1. How many wins did Notre Dame have?
2. How many wins did Notre Dame AND Baylor have?
3. How many wins did the fifth school (fifth element) have?

You have been given some code below to get you started. Fill in the code as necessary to answer these questions.


```{r bball_ex2, exercise = TRUE, eval = F, error= F, exercise.lines = 18}
# create the vector
wins_women <- c(___)

# then name the vector
names(wins_women) <- c("___", "___", "___", "___")


# How many wins did Notre Dame have?
wins_women[___]


# How many wins did Notre Dame AND Baylor have?
wins_women[c(___,___)]


# How many wins did the fifth school (fifth element) have?
wins_women[_____]


```
```{r bball_ex2-hint-1}
# create the vector
wins_women <- c(11, 8, 3, 3, 2, 2)
# then name the vector
names(wins_women) <- c("UConn", "Tennessee", "Baylor", "Stanford",
                      "South Carolina", "Notre Dame")

# How many wins did Notre Dame have?
wins_women[___]


# How many wins did Notre Dame AND Baylor have?
wins_women[c(___,___)]


# How many wins did the fifth school (fifth element) have?
wins_women[___]


```
```{r bball_ex2-hint-2}
# create the vector
wins_women <- c(11, 8, 3, 3, 2, 2)
# then name the vector
names(wins_women) <- c("UConn", "Tennessee", "Baylor", "Stanford",
                      "South Carolina", "Notre Dame")

# How many wins did Notre Dame have?
wins_women["Notre Dame"]


# How many wins did Notre Dame AND Baylor have?
wins_women[c(___,___)]


# How many wins did the fifth school (fifth element) have?
wins_women[___]


```
```{r bball_ex2-hint-3}
# create the vector
wins_women <- c(11, 8, 3, 3, 2, 2)
# then name the vector
names(wins_women) <- c("UConn", "Tennessee", "Baylor", "Stanford",
                      "South Carolina", "Notre Dame")

# How many wins did Notre Dame have?
wins_women["Notre Dame"]


# How many wins did Notre Dame AND Baylor have?
wins_women[c("Notre Dame","Baylor")]


# How many wins did the fifth school (fifth element) have?
wins_women[5]


```

## Topic 5: Indexing Using Conditional Statements

Another important skill for indexing vectors is using something called **conditional statements**. These are statements which evaluate if a certain condition is met, and returns data based on satisfying this condition.

### Introduction to conditions

Say we have a vector that gives us the ages for a set of voters:

```{r index_vec, exercise = TRUE}
#create the vector
voters <- c(18, 25, 70, 45, 47, 30)

#name the elements of the vector
names(voters) <- c("A", "B", "C", "D", "E", "F")

#return to look at the vector
voters

```

Say we want to know which voters are under the age of 40. Take a look at what happens when we run the code below:

```{r index_vec2, exercise = TRUE, exercise.setup = "index_vec"}

voters < 40

```

You can see that the output is just a list of TRUEs and FALSEs – that is, it’s a list of Boolean values. It’s TRUE for any element of the voters vector that is less than 40, and FALSE for everyone else. 

To see how this is useful, look at the following code, which puts that set of TRUEs and FALSEs into the square brackets [] you learned in the previous Topic.


```{r index_vec3, exercise = TRUE, exercise.setup = "index_vec"}

voters[voters < 40]

```

You can see this returned the same information as directly putting `c(T, T, F, F, F, T)` in the square brackets. 


### Other types of conditions

R actually has a whole set of conditions that we can use to return only the elements of a vector that meets those conditions.  

1. $==$ : is equal to
2. $!=$ : is not equal to
3. $>$ and $>=$ : greater than, greater than or equal to
4. $<$ and $<=$ : less than, less than or equal to

Let’s practice using some of these other types of conditions.

For example, if we wanted to see all voters who are exactly 25 years old, we could do this:

```{r index_vec5, exercise = TRUE, exercise.setup = "index_vec"}

voters[voters == 25]

```


Or, if we only wanted to see voters who are over the age of 80, we could do this:

```{r index_vec6, exercise = TRUE, exercise.setup = "index_vec"}

voters[voters > 80]

```
There are no voters over the age of 80!


### Practice

Now, say you have a vector that shows the test scores for a set of students:

```{r index_practice, eval = T}

#create vector
scores <- c(95, 87, 34, 75, 92)

#name elements of the vector
names(scores) <- c("A", "B", "C", "D", "E")


```

Below, write code to do the following:

1.	Index `scores` to only show the students who got a grade of 90 or higher
2.	Index `scores` to only show students who got a grade of exactly 75
3.	Index `scores`  to only show students who got a grade not equal to 34

Don’t forget to reference the list of operators above when working! The below code will give you a framework to start working with.

*1. Index scores to only show the students who got a grade of 90 or higher*

```{r index_practice1, exercise = TRUE, exercise.setup = "index_practice"}

scores[___]

```
```{r index_practice1-hint-1}

scores[scores > 90]

```

*2. Index scores to only show students who got a grade of exactly 75*

```{r index_practice2, exercise = TRUE, exercise.setup = "index_practice"}

scores[___]

```
```{r index_practice2-hint-1}

scores[scores == 75]

```

*3. Index scores  to only show students who got a grade not equal to 34*

```{r index_practice3, exercise = TRUE, exercise.setup = "index_practice"}

scores[___]

```
```{r index_practice3-hint-1}

scores[scores != 34]

```




