---
title: "Tutorial for Week 4"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)

#load in data on war casualties
war_data <- read.csv("war_casualties.csv")

war_data$casualty_est <- ifelse(war_data$war == "World War I" |
                                  war_data$war =="Italian Wars" |
                                  war_data$war == "Korean War", -99, 
                                war_data$casualty_est)


votes <- data.frame(vote_total = c(988000, 789999, 476222, 896123),
                    vote_dem = c(654333, 128093, 367806, 237809),
                    vote_rep = c(332000, 651900, 102311, 610321),
                    town = c("A", "B", "C", "D"))


```


## Welcome

Last week we introduced dataframes and started to learn how to summarize them. Critically, we only worked with variables that already had all the information we needed. In this week's class, we will continue to learn how to summarize data. But, we will be working with datasets that may have missing values for some variables, or where we need to "clean" the dataset to get the variables we want to analyze. We will also introduce a new type of variable: factors.

## Topic 1: Recoding Parts of a Variable

Sometimes, our dataset might have errors that we need to fix before we do analysis. For example, let's take a look at the variable `war` with all of the titles of the conflicts in the data set.

```{r, eval = T}
war_data$war
```

While most of the wars use fully capitalized titles, we seem to have a few that do not. In particular, it seems that World War II has lower case w's, as does the Taiping Rebellion, and the Hundred Years' War. Let's fix these so they all are properly capitalized.

Let’s start with changing “world war II” to “World War II”. To do this, we need to 1) subset to just the rows of `war_data$war` that are equal to “world war II” and then 2) redefine those cells to be capitalized. We already know how to subset to just those rows, namely


```{r, eval = T}
war_data$war[war_data$war == "world war II"]
```

Great, now we want to recode this to have proper capitalization. That is, it should read "World War II" to be uniform with the rest of the data set.

```{r}
#recode
war_data$war[war_data$war == "world war II"] <- "World War II"

#check the new value
war_data$war

```

Nice, it worked!

### Practice

Now your turn! Perform the same task as above, but now fix the "taiping rebellion" to have uniform capitalization. That is, it should read "Taiping Rebellion".

```{r recode_1, exercise = TRUE, exercise.lines = 10}

```
```{r recode_1-hint-1}
#recode
war_data$war[war_data$war == ""] <- ""

```
```{r recode_1-hint-2}
#recode
war_data$war[war_data$war == "taiping rebellion"] <- "Taiping Rebellion"

#check the new value
war_data$war
```

## Topic 2: Introducing: Missing Data

Up until now, we have largely been working with data frames that have complete observations. That is, if we have a variable for age, it has the actual age for every single row of the dataset – there are no rows where the information is missing.  Missing data is very common, and can happen for a variety of reasons. Sometimes data is hard to collect or might not exist! 

How do we handle this in our data? In R, we use a special code for missing values: `NA`, which stands for “Not Available”. This lets R know that the information is missing. Note that `NA` can appear in all types of variables, even numeric ones---R treats `NA` as its own special code. However, when you first import a new dataset, it may use different conventions for coding missing values.  Typically, you have access to clear documentation in which the creators of a data base will inform you how they coded, or labelled, missing values.

For example, it is very common for data bases to code missing values using numbers like `-99` or `-98` which are values that are very uncommon for variables to take. Let's take a look at a data set about some major wars in global history. The data set `wars_data` includes five variables:

-   `war`: the title of the conflict
-   `casualty_est`: a lower estimate of the total number of casualties
-   `start_year`: the year the conflict began
-   `end_year`: the year the conflict ended
-   `location`: where the conflict occurred

```{r, echo = T, eval = T}

summary(war_data) #summarize data


head(war_data) #view first few observations
```

When we take a look at this data, something stands out! What seems off about the `casualty_est` variable?

Well, this variable occasionally takes on the value `-99` which seems to be an uncommon value and is not a number that corresponds to a true number of causalities. We take a look at the codebook, and we see that the researchers specified that `-99` here signifies a missing value. So now we need to handle this.

If you recall from the previous topic, we can recode some part of a variable using square brackets, or `[]`. We can use this same strategy to recode `NA` values. In this case, we know we are interested in all of the cases where, for example, `casualty_est` is equal to `-99`. So, we would first index our data as such to identify these observations. It would look like this:

```{r}
war_data$casualty_est[war_data$casualty_est == -99]
```

The above line of code identifies all observations for the variable `casualty_est` within our `war_data` data set where `casualty_est == -99`. With this, we can then assign a new value to those observations. Instead of `-99`, we want these variables to have the value `NA`. We would do that like this:

```{r, eval = TRUE}
#recode variable
war_data$casualty_est[war_data$casualty_est == -99] <- NA

#summarize the variable to make sure it worked
summary(war_data$casualty_est)
```
In the summary of the `casualty_est` variable, you can now see that the minimum value is no longer `-99`, but instead is `900000`. Further, R is now reporting that there are 3 `NA` values.

Remember that `NA` is a special value in R. For it to work you must capitalize both letters. You should also never put quotes around `NA`.

## Topic 3: Summarizing with Missing Data

Missing observations can affect how we summarize our data. Many of the essential commands we use to summarize data – like `mean()` – can work even when we have missing data. But, we will have to slightly alter the way we use them. In particular, many commands take something call **optional arguments**. Before diving into the specifics of understanding how to summarize data with missing values, we first need to learn about these **optional arguments**.

### Optional Arguments in Commands

Remember from earlier in the class that **commands** (sometimes called **functions**) are directives or orders that you can give to R to perform a task. Thus far, we have used a variety of commands, such as `c()`, `mean()`, and `subset()`.

Every command accepts **arguments** or an input that is placed inside the `()`. So far, we have only worked with *required* arguments, or those that are absolutely essential for a function to work. For example, if we want to know the minimum value of a vector, we use the function `min()`. To do so, we have to give the `min()` function data:

```{r, eval = T}

new_data <- c(1, 2, 3, 5) #create a vector

min(new_data) #give the min() command our data

```

But, if we do not give `min()` any data to work with, it will not work. See here:


```{r, eval = T}

min()  #produces an error

```

We can think of functions as, for example, a recipe. Say we have a command called `hamburger()`. To make a hamburger, we at minimum have to tell the cook (in this case R) what the main ingredient is (the patty). So, we have to do as follows:

```{r, eval = F}
hamburger(beef) #for a meat lover

hamburger(veggie) #for a plant lover

```



The `hamburger()` function here takes the input -- beef or veggie -- and with its recipe, creates an output. In this case, hopefully a good hamburger.

But, functions also have **optional arguments**, just like recipes do. Let's pretend we're making hamburgers.

![](https://www.thespruceeats.com/thmb/d4-3wLGWdWQrdsYmcgOgokNDOxg=/1500x0/filters:no_upscale():max_bytes(150000):strip_icc()/vegan-tofu-veggie-burgers-recipe-3377169-hero-01-a2dd40a53b1c4d3ba21625925cc9e28b.jpg){width=250px}


When you order a hamburger at a restaurant, there are a bunch of options you can ask for. Let's start with the cook of the meat. Maybe the restaurant typically defaults to cooking a medium hamburger. But, if you want, you could say you want it well done. We would do that as follows:

```{r, eval = F}

hamburger(beef, cook = "well done")

```

Further, maybe the default for the restaurant is to give you a side of fries. But, maybe you want to opt for a potato salad today.

```{r, eval = F}
hamburger(beef, cook = "well done", side = "potato salad")
```

The important thing to note here is that *there is a typical way that the command is executed*. If you say nothing, the command will be executed in its default fashion -- in the case of a hamburger, that would be a medium hamburger with a side of french fries. However, *you have the option to change these defaults* to better fit what you want or need to do. 

Now that we understand how optional arguments in commands work, we can move on to summarizing missing data. These two topics are related as key functions, such as `mean()`, require certain optional arguments to work with data that has missing observations.

### Summarizing With Missing Data

Last week, we learned about using commands such as `mean()`, `min()`, and `max()` to summarize our data. Typically, each of these functions accept an argument `x` which is an object which contains multiple numbers. With these numbers, R would then calculate the mean, minimum (min), or maximum (max). For example, we calculate the mean of the following vector:

```{r, eval = T}

vector1 <- c(1, 5, 6, 9, 12) #create vector

mean(vector1) #calculate mean

```

But, what happens if our data has missing values? Let's see.

```{r, eval = T}

vector2 <- c(1, NA, 5, 6, 9, NA, 12) #create vector with NAs

mean(vector2) #calculate mean

```

In theory, the mean here should be the same as `vector1`, as `vector2` has the same exact numbers. However, R returns an `NA` as the mean for `vector2`. That is because R cannot calculate the mean of a an object with missing values. 

Here is where **optional arguments** come into play! Let's check out the help file for the function `mean()`. To do so, we would execute the following line in the console of RStudio.

```{r}
?mean()
```

<iframe src="https://drive.google.com/file/d/1vzCTRW9VNyUJUNvXIROpaXWqCLqE0xhG/preview" width="640" height="480" allow="autoplay"></iframe>

This help file tells us that there are two main optional arguments that we have to work with. One of them is trim, which is not relevant to us right now. But the second seems to have to do with `NA` values, or missing values. That is what we need!

This part of the help file states that within the `mean()` function, we can establish `na.rm = TRUE` or `na.rm = FALSE`. If `TRUE`, the function will remove all `NA` values and calculate the mean without them. If `FALSE` the `NA` values will stay. It also tells us that the default is `FALSE`. 

Let's see what happens if we repeat our above line of code -- where we try to calculate the mean of a vector with some `NA` values -- but now changing `na.rm` to be equal to `TRUE`.

```{r, eval = T}

vector2 <- c(1, NA, 5, 6, 9, NA, 12) #create vector with NAs

mean(vector2, na.rm = T) #calculate mean and remove NAs

```
And check it out! We no longer see that the mean is `NA`, but rather that the mean is 6.6. Specifying that the `mean()` function should then ignore all `NA` values when calculating the mean helped us with this issue!


Now, you try. Using the following code below, take `vector2` and try to calculate the mean first *without* specifying that R should remove the `NA` values. Then, try telling R to remove the RAs using the `na.rm` optional argument.

```{r meanex, exercise = TRUE}

```
```{r meanex-hint-1}
# create vector with NAs
vector2 <- c(__, NA, __, __, __, NA, __) #create vector with NAs

#calculate mean without removing NAs
mean(____)

#calculate mean and remove NAs
mean(___, na.rm = ___)
```
```{r meanex-hint-2}
# create vector with NAs
vector2 <- c(1, NA, 5, 6, 9, NA, 12) #create vector with NAs

#calculate mean without removing NAs
mean(vector2)

#calculate mean and remove NAs
mean(vector2, na.rm = TRUE)
```


Note that we can also do this with the `max()` function. Similar options are also available in other commands that we will use in the future.

## Topic 4: Creating New Variables, Part 1

In addition to recoding variables, we often want to create new variables in an existing data set. For example, let's examine a dataset which has some hypothetical information about three towns in the U.S.


```{r, eval = T}

head(votes)

```

This dataset has four variables: `town` which is a label for each town (A thru D), `vote_total` which gives us a count of all those who voted, `vote_dem` which gives us a count of how many people in the town voted for a Democrat, and `vote_rep` which gives us a count of how many people in the town voted for a Republican in a recent gubernatorial election.

Now, although we have information about how many individuals voted for the Democrat and Republican candidate, it may also be useful for us to have in our data set how many people did not vote for either (that is, they voted for a third-party or independent candidate). To do this, we will want to make a new variable called `vote_other`. 

One way to make a new variable in R is very similar to the examples of “math using vectors” that we saw in week two. Recall that if we have 

```{r, eval = T}
vec1 <- c(1, 3, 5)
vec2 <- c(2, 4, 6)
```

We can create a new vector that adds `vec1+vec2`, like so:

```{r, eval = T}
vec3 <- vec1+vec2
```

We can do something similar here. Generally, to add a new variable to a dataset, we can use the following syntax:

```{r, eval = FALSE}
dataframe$new var <- [INSTRUCTIONS FOR CREATING NEW VARIABLE]
```

In this particular case, we will need to figure out how exactly to use existing variables to create the new variable we need. To find the number of independent/third-party votes for each city, we will need to take the number of total votes, and subtract the number of Republican and Democrat votes. The below chunk of code does just that: it tells `R` to make a new variable called `vote_other`, and tells `R` how to use the `vote_total`, `vote_dem`, and `vote_rep` variables to get the answer we want. 

```{r, eval = T}

#create variable
votes$vote_other <- votes$vote_total - (votes$vote_dem + votes$vote_rep)

#check out the data
votes

```


Nice! It worked. Notice that after the `<-` operator, we have to call each variable within its data set. For example, we cannot write `vote_total - (vote_dem + vote_rep)`, but rather we must write `votes$vote_total - (votes$vote_dem + votes$vote_rep)`. You must be very specific with R to tell it from which dataset which are pulling those variables.

For example, let's check out the below error which results from only telling R to subtract `vote_dem` and `vote_rep` from `vote_total` without specifying our data set.

```{r, eval = T, error = T}

vote_total - (vote_dem + vote_rep)

```

It states that the first object it encounters, `vote_total`, is not found. That is because we have no object simply called `vote_total`, `vote_dem`, or `vote_rep`, but rather these objects only exist within our larger data set `votes`.

### Practice

Now you try! Imagine we want a new variable which captures the *percent* of all voters who voted for the Democratic candidate. We will call this new variable `dem_perc`. 

To calculate this value, you will need to divide the number of votes for the Democratic candidate by the number of total votes. You will then multiply that quotient (the value resulting from that division) by 100.


```{r dem_perc, exercise = TRUE, exercise.lines=5}

#calculate percentage


#return the full data set

```
```{r dem_perc-hint-1}

#calculate percentage
votes$dem_perc <- (votes$___ / votes$___) * ___

#return the full data set
votes

```
```{r dem_perc-hint-2}

#calculate percentage
votes$dem_perc <- (votes$vote_dem / votes$vote_total) * 100

#return the full data set
votes

```


## Topic 5: Creating New Variables, Part 2

Sometimes we can’t just use math to make a new variable. For example, let’s go back to the dataset on wars we saw earlier:

```{r, eval = T}
head(war_data)
```

Let’s say we want to create a new categorical variable that captures the severity of casualties in each war, based on `casualty_est`, the estimated number of casualties in each war. Suppose we want to classify the number of casualties as “high” or “low” depending on whether there were at least 10 million casualties. In particular, we want a new character variable that says “high” for any wars with at least 10 million casualties, and “low” for any variable with fewer than 10 million casualties.

We can do this as follows:

Step 1: Make a new variable called “casualty_category” and set it equal to `NA` for all rows.

Step 2: Select just the rows where the war had <10 million casualties, and recode our new variable to say “low” instead of `NA`. 

Step 3: Select just the rows where the war had more than 10 million casualties, and recode our new variable to say “high” instead of `NA`.

The code below does just that: 

```{r, eval = T}
#create empty new variable
war_data$casualty_category <- NA

# Create label for high casualties
war_data$casualty_category[war_data$casualty_est > 10000000] <- "high"

# Create label for low casualties
war_data$casualty_category[war_data$casualty_est <= 10000000] <- "low"

# Check out the updated data
war_data
```

Let’s take a look at what that did:

1.	We started by creating a blank “placeholder” `casualty_category` variable where all the values are NA (missing).

2.	We then took all rows where the `casualty_est` variable was greater than 7,700,000, and replaced `casualty_category` with “high”.

3.	We then took all rows where the `casualty_est` variable was less than or equal to 7,700,000, and replaced `casualty_category` with “low”.

Now it’s your turn. Below, write code to make a new variable, called `since_1800`, that is TRUE for any war that began after 1800 (that is, the `start_year` variable is at least 1800), and FALSE for any war that began prior to 1800 (that is, the `start_year` variable is prior to 1800). 

```{r since1800, exercise = TRUE, exercise.lines = 10}

```
```{r since1800-hint-1}
# create placeholder variable and set it equal to NA for all rows
war_data$___ <- ___

# create label for wars started in or after 1800
war_data$since_1800[war_data$start_year ___ 1800] <- "___"

# create label for wars started before 1800
war_data$since_1800[war_data$start_year ___ 1800] <- "___"
```
```{r since1800-hint-2}
# create placeholder variable and set it equal to NA for all rows
war_data$since_1800 <- NA

# create label for wars started in or after 1800
war_data$since_1800[war_data$start_year >= 1800] <- "TRUE"

# create label for wars started before 1800
war_data$since_1800[war_data$start_year < 1800] <- "FALSE"

# check the new value
war_data$since_1800
```

## Topic 6: Factors

So far, we’ve been working with three types of data: numeric (numbers), characters (text), and Boolean (TRUE/FALSE). When we think about data stored as characters (text), we can think about two different situations. Sometimes, the text is people’s names or other information that’s very specific to one row of the dataset. Other times, the text is telling us the category a row falls into. For example, if we have a dataset of students and a variable called “year”, that variable might categorize students into “first-year”, “sophomore”, “junior”, and “senior”. Many datasets have lots of categorical variables like this. This lesson introduces a new data type: Factors. Factors are a data type specifically designed for handling categorical data in R.

There are two main types of categorical variables: nominal and ordinal.

1. **Nominal** variables have categories with no inherent order. Examples include party affiliation (e.g., Republican, Democrat, Independent) and gender (male or female).

2. **Ordinal** variables have categories with a natural ranking or order. For example, language proficiency levels may be classified as Beginner, Intermediate, and Advanced.

### Using the factor() Function

`R` has a special type of data called a factor that is useful when we have categorical data. To see how it works, let’s first look at the limitation of using a character vector to store categorical information:



```{r eval = T}
party <- c("Republican", "Democrat", "Democrat", "Republican") #create a vector
str(party) # we see that party is currently a character variable
summary(party) # we see that str() is more informative than summary()
```

When we look at the output from `summary(party)`, we can see that it’s not very informative. Next, let’s make a new vector, `party.factor`, that stores the same information as a factor. We will do this by running the `factor()` command on the vector called `party`. 

```{r eval = T}
party.factor <- factor(party) # create a factor for party affiliation using factor()
party.factor # inspect the resulting object
```

Already, we can see a difference: when we display the party.factor vector on the screen, it shows us the values stored in the vector, but also gives us something called **levels**. In a factor variables, the list of **levels** is the set of values that variable can take. 

Now, let’s see what happens when we summarize our new factor variable:

```{r eval = T}
summary(party.factor) # now that we have converted party to a factor variable, summary() is more useful!
```

Before, `summary(party)` just told us that it was a character vector. Now, `summary(party.factor)` is giving us much more useful information: it tells us how many observations fall into each category!

### Specifying Levels

Sometimes we may want to change the order of the levels in a factor. For example, let's make Republican the first level and Democrat the second:

```{r eval = T}
party.factor2 <- factor(party, levels = c("Republican", "Democrat")) # specify the levels
str(party.factor2) # inspect output 
```

Great! Now `party.factor2` has Republican as the first level and Democrat as the second level. 

### Ordered Factors

As noted above, party affiliation is a nominal variable – there’s no natural order of categories like “Republican” and “Democrat”. Other variables are ordinal – this means they have a natural hierarchy. Let’s take the example of language proficiency levels: say an individual could be a beginner, intermediate, or advanced. We want R to store the variable so that it considers beginner “less than” intermediate, which is “less than” advanced. 

To do this, we need to add two optional inputs to the factor() command. Say we have the vector below of language levels, and we make it into an unordered factor:

```{r eval = T}
proficiency <- c("Advanced", "Beginner", "Intermediate", "Intermediate", "Advanced") #create a vector
proficiency.factor <- factor(proficiency) # create factor
str(proficiency.factor) # inspect output
```

As you can see, R puts the factors in alphabetical order, but doesn’t consider one better/worse than the other. To make it an ordered factor, we need to add two optional commands:

1.	We need to specify that it’s an ordered factor by adding `ordered=T`.
2.	We need to specify the exact order we want the levels in.

The code below remakes the vector and adds the optional commands, separated by “,”. 

```{r eval = T}
proficiency.factor <- factor(proficiency,ordered = T, levels = c("Beginner", "Intermediate", "Advanced")) # create factor
str(proficiency.factor) # inspect output
```

As you can see, R identifies `proficiency.factor` as an ordered factor with three levels: Beginner, Intermediate, and Advanced.

A lot of people find factors confusing at first. Don’t worry – we’ll spend a lot more time on this in class, including some more ways using factors, as opposed to storing data as normal text, can be useful.
