---
title: "Tutorial for Week 6"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
#library(tidyverse)
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)

# set up data for world values survey

world_values <- read.csv("wave_7_wvs.csv") %>%
  select(Q64, Q65, Q69, Q71, Q260, Q262, Q269, Q287) %>% 
  rename(conf_churches = Q64,
         conf_armforces = Q65, 
         conf_police = Q69,
         conf_govt = Q71,
         sex = Q260,
         age = Q262,
         citizen = Q269,
         class = Q287)

world_values$class <- as.factor(world_values$class)

world_values[world_values == -5] <- NA
world_values[world_values == -4] <- NA
world_values[world_values == -2] <- NA
world_values[world_values == -1] <- NA

world_values$age_group <- ifelse(world_values$age <= 25, 1,
                                 ifelse(world_values$age > 25 & world_values$age <= 35, 2,
                                        ifelse(world_values$age > 35 & world_values$age <= 45, 3,
                                               ifelse(world_values$age > 45 & world_values$age <= 55, 4,
                                                      ifelse(world_values$age > 55 & world_values$age <= 75, 5,
                                                             ifelse(world_values$age > 75, 6,
                                                                    NA))))))

world_values <- world_values %>% select(-age)

#creation of trust_byage df for plotting 
trust_byage <- world_values %>%
  select(age_group, conf_churches) %>% #select only needed variables
  filter(!is.na(age_group)) %>% #remove NAs
  group_by(age_group) %>% #group by age
  summarize(mean_trust = mean(conf_churches, na.rm = T)) #calculate the mean

```


## Welcome to Data Visualization!

This week, we will learn how to visualize our data. That is, we will learn how to examine data via graphs and figures. There are many ways to visualize data, and this often depends on the variables we are interested in. In particular, it is important to consider if the data we are plotting is discrete (aka categorical) or continuous.

The main package we will use to visualize data is called `ggplot2`, which is part of the tidyverse. Our skills in using `dplyr` and the `tidyverse` will come in handy here, as we often need to manipulate and summarize our data before visualizing it.

Note that to use `ggplot2` on your device, you will need to install and load its library. You can do so by specifically installing the `ggplot2` package and library, but also, because it is also part of the larger `tidyverse`, it is installed and loaded when you read in `tidyverse` package as well!

For example, you can do the following:


```{r, eval = F, echo = T}

library(tidyverse) #note here you do not need quotes

```

Note that you do not need to install the library for the tidyverse on your machine since we already did so last week. OR you can do the following:


```{r, eval = F, echo = T}

install.packages("ggplot2")

library(ggplot2) #note here you do not need quotes

```




## Topic 1: Understanding the Basics of Data Visualization

When visualizing data, it is important to understand that there are **many** different types of graphs and figures that you can use. We will cover a few types here.

### Histogram

We use **histograms** to plot the distribution of a variable. That is to say, the frequency of different values of that variable. For example, the below histogram from the [SAS blog](https://blogs.sas.com/content/iml/2019/03/04/visualize-sat-scores-nc.html) shows us the distribution of SAT scores in North Carolina high schools. We can see from this plot that most students in the class of 2018 got a score of about 1100. 

<iframe src="https://drive.google.com/file/d/1iEstpnb5j1T-FT267_oHs8PF5O4qoedQ/preview" width="640" height="480" allow="autoplay"></iframe>

Importantly, histograms are used for continuous data. The main variable of interest -- shown on the S axis -- is divided into "bins" or ranges of values. In this case, our main variable of interest is the SAT score. The creator of this graph created bins for each range of scores, and plotted the relative frequency of students who acheived these scores. 

### Bar Graphs

Next we have **bar graphs**, or bar charts. These are commonly confused with histograms, but they are very different. While we use histograms to plot continuous data, bar graphs are used for plotting discrete, or categorical, variables. For example, the below graphic comes from the AmericasBarometer's Assessment of [Satisfaction with Public School's in Latin America](https://www.vanderbilt.edu/lapop/insights/IO958en.pdf), and shows how the percentage of individuals satisfied with public schools varies across countries. On the X axis we have different countries (represented by their two-letter abbreviation). The Y axis instead corresponds to the percentage of individuals who reported being satisfied. 

<iframe src="https://drive.google.com/file/d/1Ncr5nNAtLroaPk6hq_xjphyWtwgR4-oH/preview" width="640" height="480" allow="autoplay"></iframe>


### Scatter Plots

**Scatter plots** help us visualize two distinct continuous variables. Each dot represents an observation, which corresponds to its values on both the X and Y axes. For example, the plot below from [The World Bank's Atlas of Sustainabile Development Goals (2023)](https://datatopics.worldbank.org/sdgatlas/goal-1-no-poverty), shows us the relationship between GDP (gross domestic prodcut) per captia (on the X axis) and the poverty rate (on the Y axis) across a variety of countries. Each dot represents a country, in this case. We can see from this plot that, in general, countries with more extreme poverty tend to have lower levels of GDP.

<iframe src="https://drive.google.com/file/d/1cGoI6PswHHi9R2SAMo_MOuK3M-1G-vcX/preview" width="640" height="400" allow="autoplay"></iframe>


### Line Graph

**Line graphs** are another means by which to plot continuous data. They are often useful to plot trends, such as trends over time or across different values of a variable. For example, the following graph comes from the [Pan American Health Organization](https://opendata.paho.org/en/core-indicators/visualizations) and examines how the total percent of the population living in urban areas in Canada and the United States has grown over time. On the X axis, we have year (a continuous variable, in this case) and on the Y axis we have the % of the population that is urban.

<iframe src="https://drive.google.com/file/d/1nMasKnkNHyDrnQQkmyDW1TzjPkotO4dY/preview" width="640" height="360" allow="autoplay"></iframe>

Importantly, this line graph also contains some categorical information -- country name! Here, color helps to capture this categorical information (a different color line for each country, with a corresponding legend). Across many types of visualizations, we can incorporate things like color or line type to help us capture categorical variables as well. So here, the authors use two different colored lines to include two distinct countries.


### Pie Chart

While there are many more types of ways to visualize data, we will conclude this introduction with **pie charts**. Pie charts are circle graphs where each slice represents a part of the whole. They help us to illustrate how variables break down in terms of their composition. For example, the below graphic comes from an analysis of race and religion in Politcs published in [The Washington Post](https://www.washingtonpost.com/news/the-fix/wp/2012/09/10/the-politics-of-race-and-religion-in-two-pie-charts/). It breaks down the religious identification and race of Republicans in 2012. It shows us, for example, that out of all Republicans included in this data, 34 percent identify as both white and evangelical Protestant. 

<iframe src="https://drive.google.com/file/d/19XJZMb_gM2kUz3j2KpcZ1iY_2L_eFa7r/preview" width="640" height="480" allow="autoplay"></iframe>


### Choosing the appropriate type of graph

Now that we have a few types of graphs under our belts, it is important to understand how we decide which type of data visualization to use. The following will help you to make this decision as you visualize your own data:

- What type of data am I plotting? Are the variables discrete or continuous?
- What relationship am I trying to show?
- What is the simplest way for me to convey this relationship?





## Topic 2: Using ggplot2

To plot data in R, we will be using the `ggplot2` package. But before we jump in there, it is essential to note that very little data is ready to be plotted from the get go. Often, we will need to summarize our data first using `dplyr` before we plot. Never forget this! 

### Plotting confidence in churches by respondent citizenship status

We're going to illustrate how to use `ggplot2` with an example. Today, we'll be using the World Values Survey, which here is saved as `world_values`. We have close to 95000 observations of individuals across the world about their opinions regarding government, trust in institutions, and their values. Let's say we are interested in examining how confidence in churches by the citizenship status of respondents.

First, run the following line of code to take a look at our data:

```{r summary1, exercise = TRUE}

summary(world_values)

```

Okay, looks like we have a few different variables here which capture confidence, in addition to respondent citizenship status, we also have sex, socioeconomic class, and the age group of respondents. For `age_group`, it is important to note that each group corresponds to the following age ranges:

* Group 1 = 16-25 years old
* Group 2 = 26-35 years old
* Group 3 = 36-45 years old
* Group 4 = 46-55 years old
* Group 5 = 56-75 years old
* Group 6 = 75 years old and above


Let's re-code `citizen` so that it is easier to understand. According to the codebook, the survey asked respondents: Are you a citizen of this country?. The value `1` corresponds to "yes I am a citizen of this country" while `2` corresponds to "no, I am not a citizen of this country."

Below, we recode this variable to have more informative information using the `ifelse()` command. Run the following line of code, and take a look at the new variable:

```{r citizen_recode, exercise = TRUE, exercise.lines = 7}

world_values$citizen <- ifelse(world_values$citizen == 1, "Citizen",
                           ifelse(world_values$citizen == 2, "Non-Citizen",
                                  NA))

#take a look at how many in each category
table(world_values$citizen)

```

Great. The codebook also tells us that for all confidence questions, higher values (in this case, a maximum of 4) corresponds to higher confidence, while lower values correspond to lower confidence.

What would be the best way to plot the relationship between these variables? Well, first we should probably consider that we need to calculate an *average level of trust* in churches for each sex. We can do so easily using dplyr. Execute the following line of code, and take a look at our new summarized data:


```{r trust_bycitizen1, exercise = TRUE, exercise.setup = "citizen_recode"}

# create new data frame called trust_bycitizen
trust_bycitizen <- world_values %>%
  select(citizen, conf_churches) %>% #select only needed variables
  filter(!is.na(citizen)) %>% #remove NAs
  group_by(citizen) %>% #group by sex
  summarize(mean_trust = mean(conf_churches, na.rm = T)) #calculate the mean trust for each group



trust_bycitizen #take a look at the final DF
```

Now we have a new data frame called `trust_bycitizen` which has the mean value of confidence in churches for both citizens and non-citizens. Since we are interested in a discrete variable (citizenship) and two average values of confidence, our best bet here is a bar graph. But how do we do that? With the `ggplot` function!

The `ggplot()` function begins as such:

```{r, eval = FALSE}

ggplot(data, aes(x = ___, y = ___)

```

Our first necessary argument is `data`. This is where we tell R what data we want `ggplot` to plot! Second, we have `aes` which stands for "aesthetics". This is where we tell R what we want to go on the X and Y axes. So, with that in mind, we would begin with the following code:

```{r, eval = FALSE}

ggplot(trust_bycitizen, aes(x = citizen, y = mean_trust))

```

Above we tell `ggplot` that our data is saved as `trust_bycitizen` and on the x axis we will be plotting `citizen` while on the y axis we will be plotting `mean_trust`. 

But, we're not done! We have not yet specified what *type* of graph we want. To specify that we want a bar graph, we will use the option`geom_col()`. We will add it on to our first line of code with a `+`. Execute the following chunk of code to see what happens:

```{r trust_bycitizen_df, echo = FALSE}
## SETUP CHUNK FOR BELOW EXERCISE

world_values$citizen <- ifelse(world_values$citizen == 1, "Citizen",
                           ifelse(world_values$citizen == 2, "Non-Citizen",
                                  NA))

# create new data frame called trust_bycitizen
trust_bycitizen <- world_values %>%
  select(citizen, conf_churches) %>% #select only needed variables
  filter(!is.na(citizen)) %>% #remove NAs
  group_by(citizen) %>% #group by sex
  summarize(mean_trust = mean(conf_churches, na.rm = T)) #calculate the mean trust for each group

```


```{r trust_plot1, exercise = TRUE, exercise.setup = "trust_bycitizen_df"}

ggplot(trust_bycitizen, aes(x = citizen, y = mean_trust)) +
  geom_col()

```

And look at that! We have a graph. Notice that non-citizens tend to have higher trust in churches.

Further, there are many different types of graphs that `ggplot` can produce. You are not expected to memorize them. Rather, we suggest that you always take a look at [this handy cheat sheet developed by Posit](https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf).



### Plotting confidence in churches by respondent age

Now let's examine a case of plotting two variables. Instead of looking at respondent sex, we are now interested in age (or in this case, age groups!). What might we do in this case?

Note that today we are using the variable `age_group`. This is to simplify things a bit, instead of using a variable which has respondents age in years. The `age_group` variable has 6 values, and we will treat it as continuous. Remember that the value of 1 correspondents to our youngest respondents, while the value of 6 corresponds to our oldest respondents.

Let's first try out a scatterplot. To do so, we will add to our `ggplot` command the line `geom_point()`, which tells the function that we are interested in plotting points. We will specify `age_group` as our X axis variable, and `conf_churches` as our Y axis variable.

```{r geom_point_test, exercise = TRUE}

ggplot(world_values, aes(x = age_group, y = conf_churches)) +
  geom_point()

```

Well, this doesn't look great. It is not a particularly informative graph. It seems like a scatterplot is not a great option here! Why might that be? Well, we only have four possible values that our y variable (confidence in churches) could take on here. So we are simply clustering all 95,000 data poitns we have on these four values. Instead, let's consider a different strategy.

Let's try a line graph. In this case, we might want to see how **average trust** varies with age. Once again, we will need to summarize our data. You will get involved for this step! Below, create a new object called `trust_byage`. Follow the same code we utilized above to summarize age by sex, but now do so with the age variable. 

```{r trust_byage, exercise = TRUE}



```
```{r trust_byage-hint-1}

# create new data frame called trust_byage
trust_byage <- ___ %>%
  select(___) %>% 
  filter(___) %>% 
  group_by(___) %>% 
  summarize(___ = mean(___, ___)) #calculate the mean


trust_byage #take a look at the final DF


```
```{r trust_byage-hint-2}

# create new data frame called trust_byage
trust_byage <- world_values %>%
  select(age_group, conf_churches) %>% #select only needed variables
  filter(!is.na(age_group)) %>% #remove NAs
  group_by(age_group) %>% #group by age
  summarize(mean_trust = mean(conf_churches, na.rm = T)) #calculate the mean

trust_byage #take a look at the final DF


```


Great! Now that we have out `trust_byage` object to work with, we can create our plot. You are going to do so using the code provided below, and with the following specifications:

- Data = `trust_byage`
- X axis = `age_group`
- Y axis = `mean_trust`
- Command for line plot = `geom_line()`



```{r trust_byage2, exercise = TRUE}

ggplot(___, aes(x = ___, y = ___)) +
  #insert type of graph here


```
```{r trust_byage2-hint-1}

ggplot(trust_byage, aes(x = age_group, y = mean_trust)) +
  geom_line()
```

This figure helps us see the relationship a bit more clearly. It seems trust in churches increases with age, until about the 5th age group (55-75), and then it begins to decrease. 



## Topic 3: Improving our visualizations

In the last topic, you learned to make basic graphs. But as-is, they are not super visually appealing, and are difficult to read. 

There are a ton of ways to make graphs prettier by adding *options*. You are not expected to memorize all of these, but here we'll go over a couple to give you a sense for how they work. 

### Axis Labels

First, it is crucial to have labels on your axes that your audience can interpret. For example, let's take our past graph about trust in the church and respondent age.

```{r}

ggplot(trust_byage, aes(x = age_group, y = mean_trust)) +
  geom_line()

```

Ideally, we would like to have the x-axis label be a bit more information. For example, "Age Group" is properly capitalized and helps our audience know the unit that we are using. To do this, we can use the `labs` argument as follows. Execute the following code chunk to check it out:

```{r  trust_byage5, exercise = TRUE, exercise.lines = 7}

ggplot(trust_byage, aes(x = age_group, y = mean_trust)) +
 geom_point() +
  geom_line() + 
  labs(x = "Age Group") # x axis label

```

Nice! It worked. Now, why don't know add a more informative label to the y axis? Started code has been provided to you below.


```{r age_plot, exercise = TRUE, exercise.lines = 7}

ggplot(trust_byage, aes(x = age_group, y = mean_trust)) +
  geom_point() +
  geom_line() +
  labs(x = "Age Group", #x axis label
       y = ___) # y axis label

```
```{r age_plot-hint-1}

ggplot(trust_byage, aes(x = age_group, y = mean_trust)) +
  geom_point() +
  geom_line() +
  labs(x = "Age Group",
       y = "Avg. Confidence in Churches (1-4 scale)")

```

Nice job!



### Plot Background

To many, a simple white background in a plot is preferable to a grey background. We can easily add this to our plot with the `theme_bw()` command. This indicates that we want the theme of our plot to be black and white (hence "bw").

Below, add a line to the plot (remember to use the `+`) and place the line `theme_bw()` at the end of the plot.

```{r age_plot2, exercise = TRUE, exercise.lines = 7}

ggplot(trust_byage, aes(x = age_group, y = mean_trust)) +
  geom_point() +
  geom_line() +
  labs(x = "Age Group", # x axis label
       y = "Avg. Confidence in Churches (1-4 scale)", # y axis label
       title = "Age vs. Confidence in Churces") + 
  ______ # add new line here with new option

```
```{r age_plot2-hint-1}

ggplot(trust_byage, aes(x = age_group, y = mean_trust)) +
  geom_point() +
  geom_line() +
  labs(x = "Age Group", # x axis label
       y = "Avg. Confidence in Churches (1-4 scale)", # y axis label
       title = "Age vs. Confidence in Churces") + # title
  theme_bw()

```



## Topic 4: Practice and Comprehension Quiz

Let's put what you learned to the test! Your task is to create two graphs. You will not have **all** of the information you need from this tutorial, and you will need to reference the [Posit ggplot2 Cheat Sheet](https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf).

Here, we will be examining trust in the government generally speaking both by respondent's citizenry (whether or not they are a citizen in their country of residence) and their socioeconomic class. The variables we will be using are as follows:

* `conf_govt`: confidence in government on a 1-4 scale 
* `citizen`: 2 = not a citizen of the country, 1 = citizen of the country
* `class`: self-reported socioeconomic class from a 1-5 scale (where 1 is lower class, and 5 is upper class)

First, let's recode `citizen` so it is a bit easier to interpret. Recode the variable so it has two values: "Citizen" and "Non-Citizen". Use an `ifelse()` statement to do so. Remember, we did this before, so this should just be a refresher!

```{r citizen, exercise = TRUE}

world_values$citizen <- ___

```
```{r citizen-hint-1}

world_values$citizen <- ifelse(___, ___,
                           ifelse(___, ___, NA))

```
```{r citizen-hint-2}

world_values$citizen <- ifelse(world_values$citizen == 1, "Citizen",
                           ifelse(world_values$citizen == 2, "Non-Citizen", NA))

```


```{r, echo = FALSE, eval = TRUE}

world_values$citizen <- ifelse(world_values$citizen == 1, "Citizen",
                           ifelse(world_values$citizen == 2, "Non-Citizen", NA))

```

**1) Bar Graph**

First, plot the average level of confidence in the government for citizens and non-citizens. Don't forget that you'll have to summarize the data first using dplyr! Note that for a bar graph, you will need to specify that `fill = "blue"` instead of color to fill the whole bar with that color. Complete the following tasks:

* Make the bars be "blue"
* Label the x axis "Citizenship Status"
* Label the y axis "Avg. Confidence in Gov't (1-4 scale)"
* Give the figure a black and white background

```{r prepare-citizen, echo = F, eval = T}
#have this here to make sure variable is properly adjusted prior to activity
world_values$citizen <- ifelse(world_values$citizen == 1, "Citizen",
                           ifelse(world_values$citizen == 2, "Non-Citizen", NA))
```


```{r citizen_govt, exercise = TRUE, exercise.setup = "prepare-citizen"}


```
```{r citizen_govt-hint-1}
citizen_govt <- ___ %>%
  select(___) %>%
  filter(___) %>%
  group_by(___) %>%
  summarize(mean_trust = ___)


ggplot(___, aes(x = ___, y = ___) +
  geom_col(___) +
  labs(___) +
  theme_bw()

```
```{r citizen_govt-hint-2}

citizen_govt <- world_values %>%
  select(conf_govt, citizen) %>%
  filter(!is.na(citizen)) %>%
  group_by(citizen) %>%
  summarize(mean_trust = mean(conf_govt, na.rm = T))


ggplot(citizen_govt, aes(x = citizen, y = mean_trust)) +
  geom_col(fill = "blue") +
  labs(x = "Citizenship Status",
       y = "Avg. Confidence in Gov't (1-4 scale)") +
  theme_bw()

```


**Challenge:** Create a similar graph (and re-summarize the data) to also include respondent class. Group by both respondent class and citizen status when summarizing trust. Plot class on the x-axis, confidence on the y-axis, and specify citizenship to be the "fill" of the bars in the `aes` argument.


```{r citizen_govt2, exercise = TRUE, exercise.setup = "prepare-citizen"}


```
```{r citizen_govt2-hint-1}

#summarize data
citizen_govt2 <- world_values %>%
  select(conf_govt, citizen, class) %>%
  filter(!is.na(___)) %>% #remove NAs for citizen variable
  filter(!is.na(___)) %>% #remove NAs for class variable
  group_by(citizen, class) %>%
  summarize(mean_trust = mean(conf_govt, na.rm = T))

#plot 
ggplot(citizen_govt2, aes(x = class, y = mean_trust, fill = citizen)) +
  geom_col(position = "dodge") +
  labs(x = "Socioeconomic Class",
       y = "Avg. Confidence in Gov't (1-4 scale)") +
  theme_bw()


```
```{r citizen_govt2-hint-2}

#summarize data
citizen_govt2 <- world_values %>%
  select(conf_govt, citizen, class) %>%
  filter(!is.na(citizen)) %>%
  filter(!is.na(class)) %>%
  group_by(citizen, class) %>%
  summarize(mean_trust = mean(conf_govt, na.rm = T))

#plot 
ggplot(citizen_govt2, aes(x = class, y = mean_trust, fill = citizen)) +
  geom_col(position = "dodge") +
  labs(x = "Socioeconomic Class",
       y = "Avg. Confidence in Gov't (1-4 scale)",
       title = "Class, Citizenship, and Confidence in Gov't") +
  theme_bw()

```

Once you have completed these tasks, please complete [this comprehension quiz.](https://unc.az1.qualtrics.com/jfe/form/SV_8wfkiUYbmaEiKfY).



