---
title: "Tutorial for Week 3"
output: 
  learnr::tutorial:
  progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(eval = FALSE)

data(mtcars)
```

## Welcome 

In this tutorial, you will learn about two overarching themes. First, we'll learn about two-dimensional R objects. We will learn about two types, called **matrices** (singular: matrix) and **data frames**. We will primarily work with data frames, as they are more flexible. Second, we will learn about how computers store files. This will give you a foundation for understanding how to import data into R.

When we discuss two-dimensional R objects, we refer to them as such because they have both rows *and* columns, while before we were working with just one-dimensional data which could be described as only a row *or* a column. For example, say we have some data about cars. A one-dimensional object may have looked just like a list of the names of the cars, like this:


```{r, echo = F, eval = T}
cars <- rownames(mtcars[1:5,])

cars
```


But now with two-dimensional data, it will take on a more complex format. For example, below we now see those five car names, and each one corresponds to a **row** in our data. Along the top, we now see names for each **column** which contain additional information about each car -- or row. For instance, we see that the Mazda Rx4 has 21 miles per gallon, while the Datsun 710 has 22.8. In this class we’ll also refer to columns as “variables”, and rows as “observations”. 

```{r, echo = F, eval = T}
head(mtcars[1:5, 1:5])
```


## Topic 1: Data Frames

Data frames are a flexible, 2D form of data. They are able to handle multiple types of data in the same object, including numbers, text, and Boolean data types. Soon, we will learn how to import data sets into R. For now, to keep things simpler, we’ll primarily work with some data sets that we’ve created for you. 

In order to use a dataset in R, you’ll first need to save it as an object. The code below creates an object for a data set that’s built into R, called `mtcars`. You can see below that each row (observation) is a different car model, and each column (variable) is a piece of information about that particular car. 

```{r save_data, exercise = TRUE}

#save mtcars data into object
mtcars <- mtcars

#look at the data set
mtcars
```


### Using a Saved Dataframe

Often, we want to look at or examine a data frame you’ve saved as an object. One easy way to do this is using the `head()` command! This will show you the first few rows of your data frame.


```{r head_demo, exercise = TRUE}

#head() data
head(mtcars)

```

Other times, we want to look at, or run commands on, a single variable (column) within a dataframe. While there are multiple ways to do this, one easy way is using the `$` notation. Here’s an example: say we want to use the `cyl` column within the `mtcars` dataframe. We could write:

```{r view_var, exercise = TRUE}

mtcars$cyl

```

This notation has a lot of other uses that you’ll learn soon. 

Finally, there are some commands that you can apply to whole dataframes to find out information like the number of rows and columns.

To ask R how many rows (or observations) our data has, we can use the command `nrow()`. This command accepts the name of the data frame, and returns the number of rows.
```{r nrow, exercise = TRUE}
nrow(mtcars) 
```

There is a similar function for the number of columns -- or variables. This is `ncol()`.

```{r ncol, exercise = TRUE}
ncol(mtcars) 
```

We can use `dim()` to ask R how many rows *and* columns a data frame has.

```{r dim, exercise = TRUE}
dim(mtcars) 
```

Now, say you want to examine that new data frame without returning the whole thing. A useful command here would be the `summary()` command. This command returns information about each of the variables. In the case of the character variables, it will tell you what class they are (character vs. numeric, for example) and the length of that variable (how many rows). In the case of numeric variables, it will give you useful information such as the minimum value, the maximum, etc. See below!

```{r summary_mtcars, exercise = TRUE}
summary(mtcars) #summarize the data

```


### Practice

Your turn! We're going to use what we just learned to explore a data set we have created for you, called `large_cities` which contains information about the top five largest cities in the United States. This data frame has four variables:

* `city`: the name of the city
* `state`: the state where that city is located
* `population`: the population of that city as of 2020
* `mayor`: the name of the mayor of that city in 2023


```{r large_cities, eval = T, echo = F}
#create vectors for the variables
city <- c("New York City", "Los Angeles", "Chicago", "Houston")
state <- c("New York", "California", "Illinois", "Texas")
population <- c(8622357, 4085014, 2670406, 2378146)
mayor <- c("Eric Adams", "Karen Bass", "Lori Lightfoot", "Sylvester Turner")


#create data frame
large_cities <- data.frame(city = city,
                             state = state,
                             population = population,
                             mayor = mayor)


```

First, take a look at the data using the `head()` and `summary()` commands:

```{r large_cities_ex1, exercise = TRUE, exercise.setup = "large_cities", exercise.lines = 6}


```
```{r large_cities_ex1-hint-1}
#head
head(large_cities)

#summary
summary(large_cities)

```


Next, determine how many columns this data set has using `ncol()` and how many observations it has using `nrow()`. Finally, put that together by using `dim()` to get the full dimensions of the data set.


```{r large_cities_ex2, exercise = TRUE, exercise.setup = "large_cities"}
#columns

#rows/observations

#dimensions

```
```{r large_cities_ex2-hint-1}
#columns
ncol(large_cities)

#rows/observations
nrow(large_cities)

#dimensions
dim(large_cities)

```




## Topic 2: Indexing Data Frames


Just like with vectors, it can be useful with data frames to select certain element(s) of our data. On the previous page, you learned that you can use the `$` notation to access a particular variable. 

This lesson teaches you to access particular rows, columns, or individual cells of a data frame. Before with vectors, we only needed to indicate the name of the element or its place within the vector using `[]`. Now, with data frames, we have to be cognizant of both the element(s) place in terms of rows AND columns.


The syntax for indexing is as follows:

`data_frame_name[row, column]`

To properly index, you MUST follow this order -- rows first, columns second. You cannot switch the order and index first by column and then row. To help you remember the order, a useful mnemonic device is to remember Roman Catholic (R always comes before C!).
As with vectors, you can index both by elements' names or place within a data frame. Again, we suggest typically using names as it might be difficult to exactly determine an element's location within the data frame. However, there are cases where both are useful.

Let's return to the previous example where we looked at the largest cities in the U.S. We will index a few elements of this data frame to demonstrate how this works. 

First, say you want to just look at a single cell of the data set - for example, the 3rd row, 2nd column. You can do so using the following code:

```{r indexing1, exercise = TRUE, exercise.setup = "large_cities"}
large_cities[3, 2]
```

Maybe instead you want to look at the whole third row. To do so, you would then leave the columns portion of the brackets blank. This tells R to show you everything in the third row. See below:

```{r indexing2, exercise = TRUE, exercise.setup = "large_cities"}
large_cities[3, ]
```

Or, what if you want to look at the full third column across all rows of data. To do so, we would leave the first space blank within our `[]`, indicating we want all rows. But, we would specify `3` for the column space, to indicate we just want the third column, like so:

```{r indexing3, exercise = TRUE, exercise.setup = "large_cities"}
large_cities[ ,3]
```

Beyond indexing by column or row **number**, we can also do so by the name of columns as well. For example, if we just want to take a look at the variable (or column) corresponding to `population`, we would do so as follows:

```{r indexing4, exercise = TRUE, exercise.setup = "large_cities"}
large_cities[ ,"population"]
```

We can also do so for multiple columns, like both `city` and `mayor`. To do so, we would put both of these columns into a `c()` command, and place them within the `[]`, like so: 


```{r indexing5, exercise = TRUE, exercise.setup = "large_cities"}

large_cities[ , c("city", "mayor")]

```

We can also combine indexing by numbers and variable names. For example, if we want to know the `mayor` but only for the first row of our data set, we would use the following code:

```{r indexing6, exercise = TRUE, exercise.setup = "large_cities"}

large_cities[1, "mayor"]

```

Further, just like we did with vectors, we can also index data frames using conditional statements. For example, if we wanted to look at cities that have a population greater than 7 million, we could do the following:

```{r indexing7, exercise = TRUE, exercise.setup = "large_cities"}

large_cities[large_cities$population > 7000000, ]

```

And notice again that we have to be specific about where within the `[]` we place this conditional statement. Here, we want to see all variables for cities that have a population of greater than 7 million. So, it is important to place this condition into the "rows" argument. We also have to use the `$` operator here to call the `population` variable before using our conditional.


So, a few things to recap here:

 * First: if you are ONLY indexing by row OR column (such as in examples 1 and 3), you leave the other corresponding space in the brackets blank. So, for example, earlier we wrote `large_cities[ , "population"]`. This effectively tells R to return ALL rows, but only the column corresponding to population
 * Second: You can combine indexing by number and row or column name.
 * Third: Just like with vectors, we can index by multiple names or numbers. Above, we used the `c()` command to return information on both city name and mayor name for all observations in our data set.
 
 
### Practice

Now you try. Below, we will use a data frame called `pres_fave` which includes information about a few presidents' favorite flowers and plants. This is fictitious information. Using this data, index the data in the following ways:

```{r pres_fave_data, echo = F, eval = T}
pres_fave <- data.frame(president = c("Obama", "Reagan", "Carter", "Kennedy",
                                      "Eisenhower", "Lincoln", "Nixon", 
                                      "Roosevelt", "Jackson"),
                        flower = c("Tulip", "Hydrangea", "Petunia", "Allium",
                                   "Crocus", "Lotus", "Petunia", "Allium",
                                   "Crocus"),
                        fruit = c("Apple", "Banana", "Apple", "Peach",
                                  "Blueberry", "Pear", "Tangerine", "Banana",
                                  "Pear"))

```

First, let's take a look at the data using `head()`:

```{r sum_presfave, exercise = TRUE, exercise.setup = "pres_fave_data"}
head(pres_fave)

```

Now, let's move on to practicing indexing. Complete the following tasks:


First, try just returning the second row of the data:

```{r pres_ex2, exercise = TRUE, exercise.setup = "pres_fave_data"}


```
```{r pres_ex2-hint-1}
pres_fave[,]
```
```{r pres_ex2-hint-2}
pres_fave[2,]
```

Now, return the value for `flower` from the 5th observation.

```{r pres_ex3, exercise = TRUE, exercise.setup = "pres_fave_data"}


```
```{r pres_ex3-hint-1}
pres_fave[,]
```
```{r pres_ex3-hint-2}
pres_fave[5,"flower"]
```

Finally, return both the `president` and `fruit` columns, but for all rows:

```{r pres_ex4, exercise = TRUE, exercise.setup = "pres_fave_data"}


```
```{r pres_ex4-hint-1}
pres_fave[,]
```
```{r pres_ex4-hint-2}
pres_fave[ , c("president", "fruit")]
```


## Topic 3: How Computers Store Files

To fully harness the power of R for data analysis and manipulation, it's important to understand how computers store files and how R interacts with them. Armed with this knowledge, you will be able to seamlessly load data sets, analyze them, and save the results! 

### The Basics

Let's start with the basics. Your computer's storage system is organized into *folders*, which can contain additional folders and *files*. These files consist of collections of data, such as text, pictures, videos, and more. Each file has a unique location on your hard drive, which we can access using something called a *path*. A file path is the address that specifies the location of a file on your computer's storage system. It's like giving directions to R on where to find the data you want to work with! 

Suppose we have the file data.csv located in a folder called "Documents." Let's look at how the file paths would appear on both Windows and macOS:

- On Windows, the file path to data.csv would be `"C:/Users/Guest/Documents/data.csv"`.

- On macOS, the file path to data.csv would be `"/Users/Guest/Documents/data.csv"`.

### Checking Your Working Directory

Finally, R has what's called a *working directory*. The working directory is the default location where R looks for files and saves outputs if no path is specified. You can check the current working directory in R using the `getwd()` function:

```{r eval = F}
getwd() #check the current working directory
```

Okay, before we move on to the next topic, take a moment to create a special folder to store all of your materials for this class. Having a dedicated POLI 281 folder will help keep your R scripts, data files, and any other class-related materials in one place, making them easier to manage and locate throughout the course!


## Topic 4: Matrices

A **martix** is another 2D type of data, however matrices are not as flexible as data frames and we will not use them very often. However, they are still important to understand! Unlike data frames, matrices can only handle data where every element is the same type (e.g., all numbers or all characters).

In R, you can create a matrix using the command `matrix()` with the following syntax.

```{r}
matrix(data, nrow, ncol, byrow, dimnames)
```

Each of these arguments is as follows:

* **data** is an input *vector* which will become all of the elements of the matrix
* **nrow** is the number of rows to be created
* **ncol** is the number of columns to be created
* **byrow** is a logical statement where `TRUE` tells the command to arrange the input data by row
* **dimname** are the names you will assign to the rows and columns

Let's check out an example! First, we will create a matrix with the numbers 1-10. We will have two rows, and we will tell R to *not* arrange the data by row:

```{r matrix1, exercise = TRUE}
my_matrix1 <- matrix(c(1:10), nrow = 2, byrow = F) #create matrix

my_matrix1 #return object
```


Now, let's see the difference if we tell R to arrange by row. What happens?

```{r  matrix2, exercise = TRUE}
my_matrix2 <- matrix(c(1:10), nrow = 2, byrow = T) #create matrix

my_matrix2 #return object
```


Above, we also told R to create two rows. What if we want 5 rows instead?

```{r matrix3, exercise = TRUE}
my_matrix3 <- matrix(c(1:10), nrow = 5, byrow = T) #create matrix

my_matrix3 #return object
```

Now you can see we will have 5 rows of data, and because we are only dealing with the numbers 1-10, we will have two columns.

### Practice

Now you try! Create a matrix called `matrix1`. Fill this matrix with numbers 1-81. The matrix should have 9 rows.

```{r matrixex_1, exercise = TRUE}

```

```{r matrixex_1-hint-1}

matrix1 <- matrix(__________) #first create the matrix

matrix1 #then return the matrix

```

```{r matrixex_1-hint-2}
matrix1 <- matrix(c(______), nrow = _____, byrow = ____) #create matrix

matrix1 #return object

```
```{r matrixex_1-hint-3}
matrix1 <- matrix(c(1:81), nrow = 9, byrow = T) #create matrix

matrix1 #return object
```


It is not so important for us to dive into this more right now, as we will not be relying on matrices too much in this class.


### Making Data Frames by Hand

Beyond making matrices from scratch, we can also make data frames ourselves as well. Remember, data frames and matrices are both 2D data objects. However, data frames can hold different types (or classes) of data in one object, while matrices can only hold one type (e.g., all integers, or all characters).

To make a data frame by hand, we will use the `data.frame()` command. The below code creates a data frame using this command. This new data frame, called `newspaper_data` contains four variables: `name` (the name of the newspaper), `n_subscribe` (the number of subscribers), and `head_q` (the location of the headquarters).

Run the following chunk of code to create the data frame using the `data.frame()` command, and then check out the data frame it created.


```{r newspaper_data, exercise = TRUE, exercise.lines = 18}

#create data frame
newspaper_data <- data.frame(name = c("USA Today", "Wall Street Journal",
                                      "The New York Times", "New York Post"),
                             #first column/variable
                             
                             n_subscribe = c(1621091, 1011200, 483701, 426129),
                             #second column/variable
                             
                             head_q = c("Virginia", "New York", "New York",
                                        "New York")
                             #third column/variable
                             
                             ) #close the command



#return the data frame
newspaper_data

```

