---
title: "Tutorial for Week 5"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
#library(tidyverse)
library(dplyr)
knitr::opts_chunk$set(echo = FALSE)

nc_congress <- read.csv("nc_congress.csv")
```

## Welcome to the Tidyverse!

So far we have been working with the commands that are built into R – this is often called “base R”. But, one advantage of R is that people can write new commands and upload them into “packages” that others can then download and use. One commonly used set of packages is called the **tidyverse**: it contains commands that will help us use and summarize data efficiently, as well as commands that make it simpler to create compelling data visualizations.

This week, we will begin working with **the tidyverse**, focusing on one of the most used packages – **`dplyr`**. This package is described by its developers as “a grammar of data manipulation” and provides to its users a consistent set of commands – called “verbs” – that help you to manipulate data. In plain language, it provides ways to streamline a lot of the tasks you’ve been learning how to do.

In addition to this tutorial and lessons from class, you may also find the online book ["R for Data Science"](https://r4ds.had.co.nz/) created by the tidyverse developers to be helpful. 

### Installing and Using Packages

When you first downloaded R, it included a bunch of built-in commands. To use commands from other packages, you will need to do two things. First, you will need to install the package: this means telling R to download the commands from the internet. You only ever do this step once. Second, you will need to load the package into R. You will need to do this in any R script where you want to use the package.

To install a package, we use the command `install.packages()`. To install on your machine, you can type the following line of code into your R console:

```{r, eval = F, echo = T}

install.packages("tidyverse") #note here that you need quotes

```

Then, to load in the package prior to using it, you will use the `library()` command, like this:


```{r, eval = F, echo = T}

library(tidyverse) #note here you do not need quotes

```

Now you’re ready to start learning the tidyverse!

## Topic 1: Important `dplyr` commands

Within `dplyr`, there are a variety of commands which help us to easily manipulate our data. Some of these are similar to those available in base R (commands available in R without installing a package). There are situations in which you may which to use one or the other. But, the advantage of using `dplyr` will become clear as we move through the lessons this week.

Some of the most useful commands from the `dplyr` package can be seen in the below table. We will walk through how we can use some of these commands, and you will also learn more about them in class this week! 

```{r, echo = F, results = 'asis'}

table <- data.frame(Command = c("select()", "filter()", "group_by()",
                                "summarise()", "mutate()", "arrange()"),
                    Description = c("Selects columns (variables)",
                                    "Filters (subsets) rows",
                                    "Groups the data",
                                    "Summarises (or aggregates) data",
                                    "Creates new variables", 
                                    "Sorts the data"))


kable(table, caption = "Important `dplyr` Commands")

```

When learning these different commands, we will be working with some data about individuals elected to represent North Carolina in the U.S.'s 118th Congress. Let's check out the data. Below, you can take a look at each of the variables and toggle through the data set. 

```{r, eval = T, echo = T}
nc_congress
```


### The `filter()` command

First, we will be using the `filter()` command, which is a command which allows us to subset our data. For example, say we only want to see information about individuals elected to the Senate in 2023. We could do the following:

```{r, eval = T, echo = T}

filter(nc_congress, branch == "Senate")

```

The first argument in the `filter()` command, as can be seen above, is the name of the data set. The second argument is the condition which we wish to satisfy for our subseting. So in this case, we asked R to only return observations where the variable `branch` was equal to "Senate".

Often, we want to save this subset of our data set into a new object. To do so, we would use the `<-` operator as before. So, if we want to save this data into a new object called `nc_senate` we would do the following:

```{r, eval = T, echo = T}

nc_senate <- filter(nc_congress, branch == "Senate")

```


### Practice 

Now you try! This time, filter for all representatives who are within the House of Representatives (labelled as "House" in the data set).

```{r house, exercise = TRUE, exercise.lines = 5}


```
```{r house-hint-1}

filter(___, branch == ___)

```
```{r house-hint-2}

filter(nc_congress, branch == "House")

```



### The `select()` command

This command allows us to subset our data set by column, or variable. That is, it allows us to only examine the variables that we wish. This is particularly useful if we have a data set with many columns, or variables, and we only need some of them.

For example, say we are just interested in the name and party of our North Carolina members of Congress, and we want to save this in a new data set called `nc_congress_parties`. We would do the following:


```{r, echo = T, eval = T}

#create new df
nc_congress_parties <- select(nc_congress, name, party)

#return df and examine
nc_congress_parties

```


We can also use the `:` operator to select multiple variables that are continguous (or next to one another). For example, say we wanted to inclue the variables `name`, `branch` and `party`. Because these variables are located next to one another in the data set, we can do the following:

```{r, echo = T, eval = T}

#select the three variables of interest
select(nc_congress, name:party)

```


### Practice

Now you try using the `select()` command. Create a new data set called `nc_districts` which includes the three variabels from the `nc_congress` data set: `name`, `branch`, and `district`.


```{r district, exercise = TRUE, exercise.lines = 5}


```
```{r district-hint-1}

nc_districts <- select(___, c(___))

```
```{r district-hint-2}

#create the new df
nc_districts <- select(nc_congress, name, branch, district)

#check out the new df
nc_districts

```



## Topic 2: Connecting dplyr commands with `%>%`

One of the main benefits of working with `dplyr` is that the package allows you to string commands together to effectively summarize your data. To do so, we use something called the pipe operator, which looks like this: `%>%`.

Within `dplyr`, recall that all verbs (or commands) take a data frame as the first argument. So, in the previous topic we wrote `filter(nc_congress, branch == "Senate"`).

If we just want to filter, that’s not too bad. But sometimes we want to use a bunch of dplyr commands in a row: maybe we want to subset, and then group our data into categories, and then calculate some summary statistics. It would be nice if we could link that all together, rather than specifying our dataframe every single time. The pipe operator `%>%` gives us a way to do that.

First, let’s look at a simple example of using `%>%`:
```{r, echo = T, eval = T}

#filter using %>%
step1 <- nc_congress %>%
  filter(branch == "House")

```

The code above starts by taking the `nc_congress` dataframe. The `%>%` tells R that whatever is on the next line of code should be applied to the `nc_congress` dataframe. If we’re just using a single `dplyr` command, that’s actually less efficient than what we did in the previous topic.

But look at the example below:

```{r, echo = T, eval = T}
with_dplyr <- nc_congress %>% #save into new object and pipe
  filter(branch == "House") %>% #filter for House and then pipe
  select(name, branch, district)
  
#check out data
with_dplyr
```


So what happened above? In the first line, we started by telling R to create a new object called `house`. To the right of the `<-` operator we put the dataset we want to work with, `nc_congress`. We finished the line with the `%>%` operator: this told R that we’re going to keep working with the `nc_congress` data on the next line. The second line of code then applies the `filter` command `nc_congress`. By putting the `%>%` operator at the end of the line, we tell R that we’re still not done! The third line takes the filtered dataset from line 2 and applies the `select` command. The result of all of this was saved into our object!

## Topic 3: Combining `group_by()` with `summarize()`

Now, we are going to introduce one of the most useful ways to use %>%: combining `group_by()` and `summarize()`. The key to understanding this topic is realizing that, if you `group_by()` and then use `%>%`, any operations on the next line of code will be ***performed separately for whatever variable you `group_by`.*** To make this more concrete, let’s look at an example.

In this dataset, the variable `years` tells us how many years a representative has been in their current position. Let’s say we want to know how Democrats and Republicans differ in terms of how long they’ve been in office. We could proceed as follows:

1. Use `group_by()` to group our data by `party` so that all subsequent calculations will be performed separately for Democrats and Republicans.

2. Use `summarize()` to create summary statistics of our data, in this case the mean of the `years` variable.

```{r, echo = T, eval = T}
mean_service <- nc_congress %>% #start with data 
  group_by(party) %>% #group by the party of each rep
  summarize(mean_service = mean(years)) #use summarize to calculate the mean

#check out the result
mean_service 
```

And what do we see in the result? We see that among those who have been elected to Congress from North Carolina, Republicans seem to have been in their seats far longer than Democrats.

If we hadn’t used `group_by` first, the `summarize()` command would have been run on the whole dataset, not broken down by party. Like this:

```{r, echo = T, eval = T}
mean_service <- nc_congress %>% #start with data
  summarize(mean_service = mean(years)) #use summarize to calculate the mean

#check out the result
mean_service
```

Next, let’s make this a bit more complicated. What if we want to know how long people have been in office, but we want to break it down by both `party` AND `branch`? We could add a second variable (argument) to the `group_by()` function:

```{r, echo = T, eval = T}
mean_service <- nc_congress %>% #start with data
  group_by(party, branch) %>% #group by the party AND branch
  summarize(mean_service = mean(years)) #use summarize to calculate the mean

#check out the result
mean_service
```

Here, we can see that particularly in the House, Republicans have served longer. We do not have any data regarding Democrat Senators, because North Carolina does not have any.

Being able to group by & then summarize is one of the big advantages of dplyr, and one that we’ll be using a lot!

## Topic 4: Less common `dplyr` commands

In addition to the previous `dplyr` commands covered, there are a variety of additional commands which are less common but still useful. These include:

* `rename()`: allows you to rename columns (variables) in your data set
* `count()`: allows you to count the observations in a specific group. Requires that you specify a grouping variable
* `tally()`: similar to count, but assumes that you have previously grouped your data using `group_by()`
* `na_if()`: useful for converting a value to `NA`

For more information on these commands, you can always visit their help files on the `dplyr` website. For example, [here is the help file for the `na_if()` command.](https://`dplyr`.tidyverse.org/reference/na_if.html).

In addition, you can always reference [this cheat sheet](https://rstudio.github.io/cheatsheets/data-transformation.pdf) for additional guidance for using `dplyr`, including information about additional commands not covered in this tutorial.


