---
title: "Tutorial for Week 7"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
#library(tidyverse)
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)


# set up data for world values survey

world_values <- read.csv("wave_7_wvs.csv") %>%
  select(Q292G, Q66, Q275, H_URBRURAL, Q76) %>% 
  rename(pol_honest = Q292G,
         conf_press = Q66, 
         edu = Q275,
         urban_rural = H_URBRURAL,
         conf_election = Q76
         )

world_values[world_values == -5] <- NA
world_values[world_values == -4] <- NA
world_values[world_values == -2] <- NA
world_values[world_values == -1] <- NA

world_values$conf_election <- ifelse(world_values$conf_election == 4, 1,
                           ifelse(world_values$conf_election == 3, 2,
                           ifelse(world_values$conf_election == 2, 3,
                           ifelse(world_values$conf_election == 1, 4, NA))))


world_values$conf_press <- ifelse(world_values$conf_press == 4, 1,
                           ifelse(world_values$conf_press == 3, 2,
                           ifelse(world_values$conf_press == 2, 3,
                           ifelse(world_values$conf_press == 1, 4, NA))))


```

## Data Visualization Part 2

Last week, we learned how to visualize data using `ggplot2`. Importantly, we also learned that in order to properly visualize data, it is often necessary to first summarize that data using `dplyr`. In this tutorial, we will discuss necessary steps to set yourself up to create good data visualizations. We will then practice creating a visualization using data from the World Values Survey.

## Steps to Good Visualizations

In order to create a good data visualization -- regardless of whether that is a bar graph, a scatterplot, or whatever you wish! -- it is important to be organized and intentional in how you approach its creation. Here, we're going to break down data visualization into six straightforward steps which will set you up to successfully illustrate relationships of interest.

To help with this, we will walk through an example. In the following section, you will then practice on your own.

### 1) Decide what relationship you want to illustrate

Before anything, you need to know what you want to visualize. What is the relationship of interest? What question are you hoping to answer?

For this example, we are interested in answering the following question: *Do people who live in urban vs. rural areas have different levels of trust in elections?*

### 2) Determine what variables are necessary for you to illustrate that relationship

Now that we have our question figured out, we need to determine what data and which variables we can use to answer the question. In this case, we are going to use data from the World Values Survey. We decide to use this data because we know that it has variables which will allow us to answer our key question. 

Let's check out the data `world_values`.

```{r}

summary(world_values)

```

Within the data set, we can see that there are two variables which should be of use to us. `urban_rural` is a variable which captures whether or not an individual lives in an urban vs. rural area. `conf_election` is a variable which captures respondents' levels of trust in elections. These variables are coded as follows:

* `urban_rural`: 1 == Urban, 2 == Rural
* `conf_election`: A scale from 1-4, where 1 == none at all and 4 == a great deal

### 3) Sketch out (yes, with pen and paper!) what you want the visualization to look like

Now that we know what variables we have, it is useful to *literally* sketch out what we are interested in visualizing. In this case, since we have one categorical variable (`urban_rural`) and one continuous variable (`conf_election`) we'll go with a bar graph. We create a little mock sketch (see below) of how we want this to look.

<iframe src="https://drive.google.com/file/d/1Gwoo5gxM61UNV1y5SDDM_v76T-ttm6A0/preview" width="480" height="450" allow="autoplay"></iframe>

Notice that we have two categories on our x-axis (for urban and rural) and the y-axis corresponds to one value of trust for each group. This provides us useful information for our next step!

### 4) Determine how you need to summarize or transform your raw data to make that sketch come to life

Currently, our data is in its raw format. Prior to creating a visual, we need to perform the appropriate transformations or summarizations of the data to allow us to create a graph.

A useful way to begin is by considering how we want our data frame itself to look. Let's consider the above sketch. It implies we need to have **two variables (or two columns)** in our data set. One should capture urban vs. rural residence, and the second should capture average confidence in elections. 

Second, we need to consider how many observations (or rows) we need in our data set. In this case, we need two data points -- an average of confidence in elections for urban citizens, and an average for rural citizens. This indicates to us that we need **two rows of data**

In all, this sounds like we need a data frame that has two columns and two rows. If we sketch it out, it should look like this:

```{r, echo = F}
library(knitr)

df <- data.frame(urban_rural = c("Rural", "Urban"), 
                 mean_conf = c("AVERAGE", "AVERAGE"))
kable(df)
```


So, how do we create this?

**First**, let's consider our `urban_rural` variable. It's current values are not particularly straightforward to follow. We need to first recode that variable so its values are more informative. We will recode values of `1` to be "Urban", values of `2` to be "Rural" and leave any other values as `NA`.

**Second,** we can now think about what we should do with the `conf_election` variable. At this point, we currently have about 95,000 respondents who each have reported a different level of trust in elections. But, we are not really interested in each of those *individual* responses, rather, we're interested in the mean level of confidence for the two groups (urban vs. rural). In that case, our best approach here would be to determine what is the *average level of trust in elections* across these two groups. 

### 5) Conduct necessary data summaries or transformations

First, let's handle the `urban_rural` variable which we discussed above.
```{r}

#use ifelse() to recode urban_rural
world_values$urban_rural <- ifelse(world_values$urban_rural == 1, "Urban",
                                   ifelse(world_values$urban_rural == 2, "Rural",
                                          NA))


```

Second, we can now summarize our data across the two groups of interest (urban vs. rural residents). We can easily summarize our data using `dplyr`, as we learned last week. In the below code, `dplyr` is used to create a new data set, titled `plot_data1`. The `world_values` data is summarized to calculate the mean level of confidence in elections for respondents who live in urban vs. rural areas. 

```{r}

plot_data1 <- world_values %>%
  na.omit() %>%
  select(urban_rural, conf_election) %>%
  group_by(urban_rural) %>%
  summarize(mean_conf = mean(conf_election, na.rm = T))

plot_data1

```


### 6) Visualize!

Now, we can plot! We will do so using the `ggplot()` command. Above, we decided that a bar graph was the best way to visualize this data. To create a bar graph, we will use the `geom_col()` argument. 

```{r}

ggplot(plot_data1, aes(x = urban_rural, y = mean_conf)) +
  geom_col()

```

A great start, but pretty basic! Let's customize this a bit with more informative axes labels, a title, and a more appealing color scheme.

```{r}

ggplot(plot_data1, aes(x = urban_rural, y = mean_conf)) +
  geom_col(fill = "skyblue2") +
  labs(x = "Urban vs. Rural Residency",
       y = "Mean Confidence in Elections (1-4 scale)",
       title = "Urban/Rural Residency and Confidence in Elections") +
  theme_bw()

```

Now, it is your turn. In the next section, you will apply these six steps to creating your own data visualization from beginning to end. 

## Practice: Part 1

Your task is to create a data visualization to answer the following question: *Do those who believe politicians are honest have higher levels of trust in the press? And does this relationship differ by education level?*

Your key variables of interest are as follows:

* `conf_press`: respondent confidence in the press (1-4 scale, 1 = none at all, 4 = a great deal)
* `pol_honest`: respondent's response to the question "Most politicians are honest and truthful, do you agree or disagree?" (1-5 scale, 5 = agree strongly, 1 = disagree strongly)
* `edu`: respondent's level of education, where:
     + 0.- Early childhood education/no education 
     + 1.- Primary education
     + 2.- Lower secondary education
     + 3.- Upper secondary education
     + 4.- Post-secondary non-tertiary education
     + 5.- Short-cycle tertiary education
     + 6.- Bachelor or equivalent
     + 7.- Master or equivalent
     + 8.- Doctoral or equivalent

Before beginning, we'll start off by summarizing the data again so you can get a look at it:

```{r}
summary(world_values)
```

There is no singular correct way to visualize this data, but we will walk through one way to do so here. Throughout this activity, we will be utilizing the six steps we just covered:

1) Decide what relationship you want to illustrate
2) Determine what variables are necessary for you to illustrate that relationship
3) Sketch out what you want the visualization to look like
4) Determine how you need to summarize or transform your raw data to make that sketch come to life
5) Conduct necessary data summaries or transformations
6) Visualize!

Let's begin!


### Step 1: Decide what relationship you want to illustrate

This has been decided for us - we want to answer the question *Do those who believe politicians are honest have higher levels of trust in the press? And does this relationship differ by education level?*

### Step 2: Determine what variables are necessary for you to illustrate that relationship

We have already been provided with a data set which has three key variables that will allow us to illustrate the relationship of interest. These variables are `conf_press`, `pol_honest`, and `edu`.

### Step 3: Sketch out what you want the visualization to look like

On your own, with a pen and paper, sketch out how you might imagine this data visualization should look. Consider if the variables you are interested in are continuous or categorical. On the next page, you will be provided with a sketch which you will then use to inform your data visualization.


## Practice: Part 2

Since we cannot see the sketch that you came up with, we will use the below sketch to dictate how we want to visualize our data moving forward.

<iframe src="https://drive.google.com/file/d/1P2jYgEEix13WI9LFfZKZs5Lje9vlcwpj/preview" width="390" height="480" allow="autoplay"></iframe>

### Step 4: Determine how you need to summarize or transform your raw data to make that sketch come to life

So, how do we transform our data to make this visualization come to life? What do we need our data to look like?

On the x-axis, we have honesty, which is a 5-point scale. On the y-axis, we have average confidence. And then we are also interested in breaking this up by two groups of education levels (higher vs. lower levels of education). So, we will want two averages for each level of honest -- 10 observations or rows! And then, we have three variables or columns. We can think about our data looking like this:

```{r, echo = F}
df2 <- data.frame(pol_honest = c(1, 1, 2, 2, 3, 3, 4, 4, 5, 5), 
                 edu = c("High", "Low", "High", "Low", "High", "Low",
                               "High", "Low", "High", "Low"),
                 mean_conf = rep("AVERAGE", length.out = 10))
kable(df2)
```

From looking at this sketch of the data, it seems we will need to do the following:

1) Create a categorical variable for education
2) Calculate average levels of confidence grouped by both education and ratings of politician honesty

### Step 5: Conduct necessary data summaries or transformations

First, we'll need to make our education variable categorical. Below, create a new variable called `ed_cat`. This variable should == "High School and Below" for all educational categories from 0-3, and should == "Post-Secondary and Above" for all educational categories from 4-8. Make this variable into a factor variable.

```{r practice1, exercise = TRUE}


```
```{r practice1-hint-1}

#transformation edu to be ed_cat using ifelse()
world_values$ed_cat <- factor(ifelse(world_values$edu <= ___, ___,
                              ifelse(world_values$edu > ___, ___,
                                    ___)))

#check it out to make sure it worked
summary(___)
```
```{r practice1-hint-2}

#transformation edu to be ed_cat using ifelse()
world_values$ed_cat <- factor(ifelse(world_values$edu <= 3, "High School and Below",
                              ifelse(world_values$edu > 3, "Post-Secondary and Above",
                                     NA)))

#check it out to make sure it worked
summary(world_values$ed_cat)
```


Second, we will want to summarizing our data to calculate average levels of confidence grouped by both education and ratings of politician honesty.

To do so, we will use `dplyr`. Below, create a new data frame called `plot_data` that summarizes the `world_values` data. You will want to create a new variable `mean_conf` with the average level of confidence in the press, grouped by both `pol_honest` and `ed_cat`. The below code should get you started!

```{r practice2_setup, echo = F}
#transformation edu to be ed_cat using ifelse()
world_values$ed_cat <- factor(ifelse(world_values$edu <= 3, "High School and Below",
                              ifelse(world_values$edu > 3, "Post-Secondary and Above",
                                     NA)))

```
```{r practice2, exercise = TRUE, exercise.setup = "practice2_setup"}

___ <- ___ %>% #create new df
  ___() %>% # remove NA values
group_by(___, ___) %>% # group by for summarizing
summarize(___ = mean(___))


#return your data
___
```
```{r practice2-hint-1}

plot_data <- world_values %>% #create new df
  na.omit() %>% # remove NA values
group_by(pol_honest, ed_cat) %>% # group by for summarizing
summarize(mean_conf = mean(conf_press))


#return your data
plot_data
```


### Step 6: Visualize!

Now that we have our data properly summarized, we can create a visualization using `ggplot2`. We'll first begin with the basics, and then work on getting the figure to look aesthetically pleasing.

```{r visual_setup, echo = F, exercise.setup = "practice2_setup"}

plot_data <- world_values %>% #create new df
  na.omit() %>% # remove NA values
group_by(pol_honest, ed_cat) %>% # group by for summarizing
summarize(mean_conf = mean(conf_press))

```
```{r visual_prac, exercise = TRUE, exercise.setup = "visual_setup", exercise.lines = 8}

ggplot(___, aes(x = ___, y = ___, color = ___)) + # begin plot with x, y axis and color 
  ___ + #include dots
  ___  #include lines
  

```
```{r visual_prac-hint-1}

ggplot(plot_data, aes(x = pol_honest, y = mean_conf, color = ed_cat)) + # begin plot with
  # x and y axis, also specify color for education
  geom_point() + #include dots
  geom_line()  #include line


```

Great. Now that we have the basics established, let's work on making this look a little more visually pleasing. Below, add the following elements to the plot:

* A black and white theme
* X and y axes labels
* A title
* A label for the legend (color)

```{r visual_prac2, exercise = TRUE, exercise.setup = "visual_setup", exercise.lines = 10}

ggplot(plot_data, aes(x = pol_honest, y = mean_conf, color = ed_cat)) + # begin plot with
  # x and y axis, also specify color for education
  geom_point() + #include dots
  geom_line() + #include line
  ___ + # black and white theme
  labs(___, # x axis label
       ___, # y axis label
       ___, #title
       ___) #legend title

```
```{r visual_prac2-hint-1}

ggplot(plot_data, aes(x = pol_honest, y = mean_conf, color = ed_cat)) + # begin plot with
  # x and y axis, also specify color for education
  geom_point() + #include dots
  geom_line() + #include line
  theme_bw() + # black and white theme
  labs(x = "Politicians Are Honest (1-5)", # x axis label
       y = "Avg. Confidence in Press (1-4 scale)", # y axis label
       title = "Perception of Politician Honesty vs. Confidence in Press", #title
       color = "Education") #legend title

```



## Additional Options

In addition, there are other ways that we could have plotted the data. As mentioned before, there is often not only one correct way to visualize data. Below, we have included some additional plots that we could have created, based on our initial sketch and goals for our end visualization. 

```{r, echo = F}

#transformation edu to be ed_cat using ifelse()
world_values$ed_cat <- ifelse(world_values$edu <= 3, "High School and Below",
                              ifelse(world_values$edu > 3, "Post-Secondary and Above",
                                     NA))
```


### Alternative 1

Here, instead of using colors to distinguish between levels of education, we do something called "faceting", which separates the figure into two panels.
```{r}

world_values %>%
  na.omit() %>% # remove NA values
group_by(pol_honest, ed_cat) %>% # group by for summarizing
summarize(mean_conf = mean(conf_press)) %>% # calculate mean level of confidence
  ggplot(aes(x = pol_honest, y = mean_conf)) + # begin plot with x and y axis
  geom_point() + #include dots
  geom_line() + #include line
  facet_wrap(~ed_cat) + #facet wrap by education
  theme_bw() + # black and white theme
  labs(x = "Politicians Are Honest (1-5)", # x axis label
       y = "Avg. Confidence in Press (1-4 scale)", # y axis label
       title = "Perception of Politician Honesty vs. Confidence in Press", #title
       color = "Education") #legend title
```

### Alternative 2

Here, we decided to also make our `pol_honest` variable categorical. Since one of our main variables is now categorical, we went with a bar graph! We calculated the average confidence in the press by education level, with different colored bars for each category of education. 

```{r}
#transformation pol_honest to be honest_cat using ifelse()
world_values$honest_cat <- ifelse(world_values$pol_honest <= 3, "Disagree",
                              ifelse(world_values$pol_honest > 3, "Agree",
                                     NA))

world_values %>%
  na.omit() %>% #remove NAs
group_by(honest_cat, ed_cat) %>% #group by for summarizing
summarize(mean_conf = mean(conf_press)) %>% #calculate mean confidence
  ggplot(aes(x = honest_cat, y = mean_conf, fill = ed_cat)) + # specify our
  # x and y axes, and also that we will have different fill colors by edu
  geom_col(position = "dodge") + #bar graph, with "dodged" bars (side by side)
  theme_bw() +  # black and white theme
  labs(x = "Politicians Are Honest (1-5)", # x axis label
       y = "Avg. Confidence in Press (1-4 scale)", # y axis label
       title = "Perception of Politician Honesty vs. Confidence in Press", #title
       fill = "Education") #legend title


```



