---
title: "Tutorial for Week 7"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
#library(tidyverse)
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)


# set up data for world values survey

world_values <- read.csv("wave_7_wvs.csv") %>%
  select(Q292G, Q66, Q275, H_URBRURAL, Q76) %>% 
  rename(pol_honest = Q292G,
         edu = Q275,
         urban_rural = H_URBRURAL,
         conf_election = Q76
         )

world_values[world_values == -5] <- NA
world_values[world_values == -4] <- NA
world_values[world_values == -2] <- NA
world_values[world_values == -1] <- NA

world_values$conf_election <- ifelse(world_values$conf_election == 4, 1,
                           ifelse(world_values$conf_election == 3, 2,
                           ifelse(world_values$conf_election == 2, 3,
                           ifelse(world_values$conf_election == 1, 4, NA))))
```

## Welcome!

Last week, we learned how to visualize data using `ggplot2`. Importantly, we also learned that in order to properly visualize data, it is often necessary to first summarize that data using `dplyr`. In this tutorial, we will discuss necessary steps to set yourself up to create good data visualizations. We will then practice creating a visualization using data from the World Values Survey.

## Topic 1: Steps to Good Visualizations

In order to create a good data visualization -- regardless of whether that is a bar graph, a scatterplot, or whatever you wish! -- it is important to be organized and intentional in how you approach its creation. Here, we're going to break down data visualization into six straightforward steps which will set you up to successfully illustrate relationships of interest.

To help with this, we will walk through an example. In the following section, you will then practice on your own.

### 1) Decide what relationship you want to illustrate

Before anything, you need to know what you want to visualize. What is the relationship of interest? What question are you hoping to answer?

For this example, we are interested in answering the following question: *Do people who live in urban vs. rural areas have different levels of trust in elections?*

### 2) Determine what variables are necessary for you to illustrate that relationship

Now that we have our question figured out, we need to determine what data and which variables we can use to answer the question. In this case, we are going to use data from the World Values Survey. We decide to use this data because we know that it has variables which will allow us to answer our key question. 

Let's check out the data `world_values`.

```{r}

summary(world_values)

```

Within the data set, we can see that there are two variables which should be of use to us. `urban_rural` is a variable which captures whether or not an individual lives in an urban vs. rural area. `conf_election` is a variable which captures respondents' levels of trust in elections. These variables are coded as follows:

* `urban_rural`: 1 == Urban, 2 == Rural
* `conf_election`: A scale from 1-4, where 1 == none at all and 4 == a great deal

### 3) Sketch out (yes, with pen and paper!) what you want the visualization to look like

Now that we know what variables we have, we need to think about what the graph should look like. Here you should think about:

1. Which variable should go on the x axis? Y axis?
2. What type of graph should you use here?

First, consider the axes. In this case, our research question is “Do people who live in urban vs. rural areas have different levels of trust in elections?” Generally, we want to put our outcome of interest on the y axis. Here, that’s “trust in elections”, measured by `conf_election`. And we want to show that variable broken down by whether someone is urban or rural. So `urban_rural` will go on our x axis.

It is useful to *literally* sketch out what we are interested in visualizing. In this case, since our x axis (`urban_rural`) is a categorical variable, and we’ll be summarizing trust, a bar graph would make sense. Here’s a  little mock sketch (see below) of how we want this to look.

<iframe src="https://drive.google.com/file/d/1MUkPln0MEzjWFb0RhRQRUn4pWdXNJaWs/preview" width="640" height="480" allow="autoplay"></iframe>

Notice that we have two categories on our x-axis (for urban and rural) and the y-axis corresponds to one value of trust for each group. This provides us useful information for our next step!

### 4) Determine how you need to summarize or transform your raw data to make that sketch come to life

Currently, our data is in its raw format. Prior to creating a visual, we need to perform the appropriate transformations or summarizations of the data to allow us to create a graph.

A useful way to begin is by considering how we want our data frame itself to look. Let's consider the above sketch. It implies we need to have **two variables (or two columns)** in our data set. One should capture urban vs. rural residence, and the second should capture average confidence in elections. 

Second, we need to consider how many observations (or rows) we need in our data set. In this case, we need two data points -- an average of confidence in elections for urban citizens, and an average for rural citizens. This indicates to us that we need **two rows of data**

In all, this sounds like we need a data frame that has two columns and two rows. If we sketch it out, it should look like this:

```{r, echo = F}
library(knitr)

df <- data.frame(urban_rural = c("Rural", "Urban"), 
                 mean_conf = c("AVERAGE", "AVERAGE"))
kable(df)
```


So, how do we create this?

**First,** let’s consider our urban_rural variable. It’s currently coded so that `1` means “Urban” and `2` means “Rural. We should probably recode that to be more informative. We will make a new character variable called “urban” that is coded as “Urban” or “Rural”, leaving any current NAs. 

**Second,** we can now think about what we should do with the `conf_election` variable. At this point, we currently have about 95,000 respondents who each have reported a different level of trust in elections. But, we are not really interested in each of those *individual* responses, rather, we're interested in the mean level of confidence for the two groups (urban vs. rural). In that case, our best approach here would be to determine what is the *average level of trust in elections* across these two groups. 

### 5) Conduct necessary data summaries or transformations

First, let's handle the `urban_rural` variable which we discussed above.
```{r}
# create a new variable called urban
world_values$urban <- "NA"
world_values$urban[world_values$urban_rural == 1] <- "Urban"
world_values$urban[world_values$urban_rural == 2] <- "Rural"
```

Second, we can now summarize our data across the two groups of interest (urban vs. rural residents). We can easily summarize our data using `dplyr`, as we learned last week. In the below code, `dplyr` is used to create a new data set, titled `plot_data1`. The `world_values` data is summarized to calculate the mean level of confidence in elections for respondents who live in urban vs. rural areas.

Note that we’re using a new `dplyr` “verb” here: `na.omit()`. This will drop rows with NAs, so that  we don’t end up with a third bar for people who are NA on the urban variable. 

```{r}

plot_data1 <- world_values %>%
  na.omit() %>%
  group_by(urban) %>%
  summarize(mean_conf = mean(conf_election, na.rm = T))

plot_data1
```


### 6) Visualize!

Now, we can plot! We will do so using the `ggplot()` command. Above, we decided that a bar graph was the best way to visualize this data. To create a bar graph, we will use the `geom_col()` argument. 

```{r}

ggplot(plot_data1, aes(x = urban, y = mean_conf)) +
  geom_col()

```

This is a great start, but it's pretty basic! Let's customize this a bit with more informative axes labels, a title, and a more appealing color scheme.

```{r}

ggplot(plot_data1, aes(x = urban, y = mean_conf)) +
  geom_col(fill = "skyblue2") +
  labs(x = "Urban vs. Rural Residency",
       y = "Mean Confidence in Elections (1-4 scale)",
       title = "Urban/Rural Residency and Confidence in Elections") +
  theme_bw()

```

Now, it is your turn. In the next section, you will apply these six steps to creating your own data visualization from beginning to end. 

## Topic 2: Practice (Part 1)

Your task is to create a data visualization to answer the following question: *Are people who are more educated more likely to believe politicians are honest?*

Your key variables of interest are as follows:

* `pol_honest`: respondent's response to the question "Most politicians are honest and truthful, do you agree or disagree?" (1-5 scale, 5 = agree strongly, 1 = disagree strongly)
* `edu`: respondent's level of education, where:
     + 0.- Early childhood education/no education 
     + 1.- Primary education
     + 2.- Lower secondary education
     + 3.- Upper secondary education
     + 4.- Post-secondary non-tertiary education
     + 5.- Short-cycle tertiary education
     + 6.- Bachelor or equivalent
     + 7.- Master or equivalent
     + 8.- Doctoral or equivalent

Before beginning, we'll start off by summarizing the data again so you can get a look at it:

```{r}
summary(world_values)
```

There is no singular correct way to visualize this data, but we will walk through one way to do so here. Throughout this activity, we will be utilizing the six steps we just covered:

1) Decide what relationship you want to illustrate
2) Determine what variables are necessary for you to illustrate that relationship
3) Sketch out what you want the visualization to look like
4) Determine how you need to summarize or transform your raw data to make that sketch come to life
5) Conduct necessary data summaries or transformations
6) Visualize!

Let's begin!


### Step 1: Decide what relationship you want to illustrate

This has been decided for us - we want to answer the question *Are people who are more educated more likely to believe politicians are honest?*

### Step 2: Determine what variables are necessary for you to illustrate that relationship

We have already been provided with a data set which has the key variables that will allow us to illustrate the relationship of interest. These variables are `pol_honest` and `edu`.

### Step 3: Sketch out what you want the visualization to look like

Now, on your own, with a pen and paper, sketch out how you might imagine this data visualization should look. Remember you’ll need to think about:

1. What variable should go on the x axis? On the y axis?
2. Are these variables continuous or categorical?
3. What type of graph do you want to make? Bar graph? Line graph? Something else? 

Do your best to sketch out a graph, but don’t worry if you get stuck -- on the next page, you will be provided with a sketch which you will then use to inform your data visualization.


## Topic 3: Practice (Part 2)

Since we cannot see the sketch that you came up with, we will use the below sketch to dictate how we want to visualize our data moving forward.

<iframe src="https://drive.google.com/file/d/1-I9uTI_bqpGgS1kzRVxIW17dBd9cMJby/preview" width="640" height="480" allow="autoplay"></iframe>

### Step 4: Determine how you need to summarize or transform your raw data to make that sketch come to life

So, how do we transform our data to make this visualization come to life? What do we need our data to look like?

In our bar graph sketch, we had every education level on the x-axis. Recall from the previous page that education can take on numeric values from 0 (`no education`) to 8 (doctorate or equivalent). On the y-axis, we had average belief in politician honesty (`pol_honest`), which is a 5-point scale where 1 = strongly disagree politicians are honest and 5 = strongly agree politicians are honest.This suggests that we’ll want a summary dataset with 2 columns: one for different levels of education, and one for average political honesty. Because `edu` can take on 9 values, we’ll need 9 rows in our summary dataset, which should look something like this:


```{r, echo = F}
df2 <- data.frame( edu = c("0",
                         "1",
                         "2",
                         "3",
                         "4",
                         "5",
                         "6",
                         "7",
                         "8"), pol_honest = c("Average", "Average", "Average", "Average", "Average", "Average", "Average", "Average", "Average"))
kable(df2)
```

From looking at this sketch of the data, it seems we will need to do the following:

1.	Make a version of the education variable where we replace the numbers 0-8 with more meaningful labels.

2.	Calculate average beliefs that politicians are honest for each level of education.

### Step 5: Conduct necessary data summaries or transformations

First, we’ll need to make our education variable. Run the code below to turn edu into a factor variable called `ed_cat` where we define the labels:

```{r, echo = T}
# create levels and labels
levels_edu <- c(0, 1, 2, 3, 4, 5, 6, 7, 8)
labels_edu <- c(
  "Early childhood education/no education",
  "Primary education",
  "Lower secondary education",
  "Upper secondary education",
  "Post-secondary non-tertiary education",
  "Short-cycle tertiary education",
  "Bachelor or equivalent",
  "Master or equivalent",
  "Doctoral or equivalent"
)

# turn into factor
world_values$ed_cat <- factor(world_values$edu, levels = levels_edu, labels = labels_edu)

# examine output
str(world_values$ed_cat)
```

Now, we can make our summary dataset using `dplyr`. Below, create a new data frame called `plot_data` that summarizes the `world_values` data. You will want to create a new variable `mean_honest` with the average level of `pol_honest`, grouped by `ed_cat`. The below code should get you started!

```{r summary1_setup, echo=F}
# create levels and labels
levels_edu <- c(0, 1, 2, 3, 4, 5, 6, 7, 8)
labels_edu <- c(
  "Early childhood education/no education",
  "Primary education",
  "Lower secondary education",
  "Upper secondary education",
  "Post-secondary non-tertiary education",
  "Short-cycle tertiary education",
  "Bachelor or equivalent",
  "Master or equivalent",
  "Doctoral or equivalent"
)

# convert to factor
world_values$ed_cat <- factor(world_values$edu, levels = levels_edu, labels = labels_edu)
```
```{r summary1, exercise = TRUE, exercise.setup = "summary1_setup"}
# summarize data
____ <- _____ %>% # create new dataframe 
  ____ %>% # remove NA values
  group_by(____) %>% # group by for summarizing 
  summarize(____ = mean(____)) 

# return your data
head(_____)
```
```{r summary1-hint-1}
# summarize data
plot_data <- world_values %>% #create new df
  na.omit() %>% # remove NA values
group_by(ed_cat) %>% # group by for summarizing
summarize(mean_honest = mean(pol_honest))

#return your data
plot_data
```

### Step 6: Visualize!

Now that we have our data properly summarized, we can create a visualization using `ggplot2`. We’ll first begin with the basics, and then work on getting the figure to look aesthetically pleasing. Fill in the code template below to make the graph!

```{r visual_setup, echo = F}

# create levels and labels
levels_edu <- c(0, 1, 2, 3, 4, 5, 6, 7, 8)
labels_edu <- c(
  "Early childhood education/no education",
  "Primary education",
  "Lower secondary education",
  "Upper secondary education",
  "Post-secondary non-tertiary education",
  "Short-cycle tertiary education",
  "Bachelor or equivalent",
  "Master or equivalent",
  "Doctoral or equivalent"
)

# convert to factor
world_values$ed_cat <- factor(world_values$edu, levels = levels_edu, labels = labels_edu)

plot_data <- world_values %>% #create new df
  na.omit() %>% # remove NA values
group_by(ed_cat) %>% # group by for summarizing
summarize(mean_honest = mean(pol_honest))

```
```{r visual_prac, exercise = TRUE, exercise.setup = "visual_setup", exercise.lines = 8}

ggplot(___, aes(x = ___, y = ___)) + # begin plot with x, y axis and color 
  _____ # graph type  
  

```
```{r visual_prac-hint-1}

ggplot(plot_data, aes(x = ed_cat, y = mean_honest)) + # begin plot with
  # x and y axis
  geom_col()
```

Great. Now that we have the basics established, let’s work on making this look a little more visually pleasing. Below, add the following elements to the plot:

 * A black and white theme
 * X and y axes labels
 * A title
 * Challenge: Change the angle of the education categories to be at 90 degrees 


```{r visual_prac2, exercise = TRUE, exercise.setup = "visual_setup", exercise.lines = 10}

ggplot(plot_data, aes(x = ed_cat, y = mean_honest)) + # begin plot with
  # x and y axis
  geom_col() +
  ___ + # black and white theme
  labs(___, # x axis label
       ___, # y axis label
       ___, #title
       ___) #legend title

```
```{r visual_prac2-hint-1}


ggplot(plot_data, aes(x = ed_cat, y = mean_honest)) + # begin plot with
  # x and y axis
  geom_col() +
  theme_bw() + # black and white theme
  labs(x = "Level of Education", # x axis label
       y = "Mean Politician Honesty", # y axis label
       title = "Mean Perception of Politician Honesty vs. Education") + #title
  theme(axis.text.x=element_text(angle=90))

```

## Topic 4: Additional Options

In addition, there are other ways that we could have plotted the data. As mentioned before, there is often not only one correct way to visualize data. Below, we have included some additional plots that we could have created, based on our initial sketch and goals for our end visualization. 

### Alternative 1

One option is to plot a line graph instead of a bar graph. Here is the code to do just that. Note that we use `geom_line()` instead of `geom_col()`. We also use `geom_point(color="red")` to draw attention to the individual data points.

```{r alt1_setup, echo = F}

# create levels and labels
levels_edu <- c(0, 1, 2, 3, 4, 5, 6, 7, 8)
labels_edu <- c(
  "Early childhood education/no education",
  "Primary education",
  "Lower secondary education",
  "Upper secondary education",
  "Post-secondary non-tertiary education",
  "Short-cycle tertiary education",
  "Bachelor or equivalent",
  "Master or equivalent",
  "Doctoral or equivalent"
)

# convert to factor
world_values$ed_cat <- factor(world_values$edu, levels = levels_edu, labels = labels_edu)

plot_data <- world_values %>% #create new df
  na.omit() %>% # remove NA values
group_by(ed_cat) %>% # group by for summarizing
summarize(mean_honest = mean(pol_honest))
```

```{r}
ggplot(plot_data, aes(x = ed_cat, y = mean_honest)) + # begin plot with x and y axis
  geom_line(aes(group=1), color="blue") + # use geom_line() instead of geom_col()
  geom_point(color="red") + # use geom_point() to draw attention to the data points
  theme_bw() + # black and white theme
  labs(x = "Level of Education", # x axis label
       y = "Mean Politician Honesty", # y axis label
       title = "Mean Perception of Politician Honesty vs. Education") + #title
  theme(axis.text.x=element_text(angle=90))
```

### Alternative 2

Another option would be to collapse our education variable into three levels (low, medium, and high) and examine how perceptions of honesty vary across the levels. To do this, we need to recode our education variable and create a new data frame that calculates the mean perception of honesty for each level. The code below does just that.

```{r alt2_setup, echo = FALSE}
# create levels and labels
levels_edu <- c(0, 1, 2, 3, 4, 5, 6, 7, 8)
labels_edu <- c(
  "Early childhood education/no education",
  "Primary education",
  "Lower secondary education",
  "Upper secondary education",
  "Post-secondary non-tertiary education",
  "Short-cycle tertiary education",
  "Bachelor or equivalent",
  "Master or equivalent",
  "Doctoral or equivalent"
)

# convert to factor
world_values$ed_cat <- factor(world_values$edu, levels = levels_edu, labels = labels_edu)
```

```{r}
# break the education variable into three levels: low, medium, high
world_values$new_ed_cat <- NA  # create the new column and fill with NAs

world_values$new_ed_cat[world_values$edu <= 3] <- "Low"
world_values$new_ed_cat[world_values$edu >= 4 & world_values$edu <= 6] <- "Medium"
world_values$new_ed_cat[world_values$edu >= 7] <- "High"

# transform new_ed_cat variable into a factor and specify levels
world_values$new_ed_cat <- factor(world_values$new_ed_cat, levels = c("Low", "Medium", "High"))

# create a new dataframe for plotting
plot_data2 <- world_values %>%
  na.omit() %>% # eliminate NAs
  group_by(new_ed_cat) %>% # group by new_ed_cat
  summarize(new_pol_honest = mean(pol_honest, na.rm = TRUE)) # summarize new_pol_honest

# take a look at the data
print(plot_data2)
```

We are now ready to plot. We can do this using the following code:

```{r}
ggplot(plot_data2, aes(x = new_ed_cat, y = new_pol_honest)) + 
  geom_col(fill="skyblue") + 
  theme_bw() + 
  labs(x = "Level of Education",
       y = "Mean Politician Honesty",
       title = "Mean Perception of Politician Honesty vs. Education") + 
  theme(axis.text.x=element_text(angle=90))
```

Interesting! We see that those in the "Low" category see politicians as *slightly* more honest (mean = 2.82) than those in the "Medium" (mean = 2.63) and "High" (mean = 2.80) categories.